---
title: Bumblebee probability of occurrence responds to complex interactions between local and landscape land use, climatic niche properties and climate change
author: "Tim Newbold^1,\\*^, Jeremy Kerr^2^, Peter Soroye^2,&dagger;^, Jessica J. Williams^1^"
output:
  bookdown::word_document2:
    toc: false
    fig_caption: TRUE
    number_sections: no
always_allow_html: true
csl: ecology-letters.csl
bibliography: bumblebees_lu_cc_refs.bib
---

^1^Centre for Biodiversity and Environment Research, Department of Genetics, Evolution and Environment, University College London, London, UK

^2^Department of Biology, University of Ottawa, Ottawa, Ontario, Canada

^&dagger;^Current address: Wildlife Conservation Society Canada, Toronto, Ontario, Canada

^\*^Correspondence: Tim Newbold, Centre for Biodiversity & Environment Research, Department of Genetics, Evolution & Environment, University College London, Gower Street, London WC1E 6BT, United Kingdom; Email: [t.newbold@ucl.ac.uk](mailto:t.newbold@ucl.ac.uk); Tel: +44(0)2031087569

**Short title:** Land-use-climate interactions and bumblebees

**Keywords:** Bumblebees, Climate change, Fertilizer, Land use, Natural habitat, Occupancy, Pesticide, Pressure interactions, Thermal niche

**Type of article:** Letter

**Authorship statement:** TN, JK, PS and JJW designed the analyses. TN performed all main analyses. TN and PS provided the methods for quantifying species' thermal niche position. TN wrote the first draft of the manuscript. All authors contributed to editing the manuscript. TN and JK secured the funding to support the presented research.

**Number of words in abstract:** 150

**Number of words in main text (excluding abstract, acknowledgments, references, and table and figure legends):** 4,986

**Number of references:** 78

**Number of figures:** 5

**Number of tables:** 1

<!-- SETUP: Global options -->
```{r setup,echo=FALSE,results=FALSE}
knitr::opts_chunk$set
options(scipen=20)
```

\newpage

# Abstract

Insect biodiversity is changing rapidly, driven by a complex suite of pressures, notably land use, land-use intensification, and increasingly climate change. Bumblebees pollinate wild plants and crops, but we lack large-scale evidence on how land use and climate change interact to drive bumblebee biodiversity changes. We assess bumblebee responses to interactive effects of land use and climate pressures across North America and Western Europe. Occupancy increases with landscape natural habitat and in landscapes more recently disturbed by humans. Responses to historical climate warming are negative in natural habitats but positive in human land uses, while human land use reduces occupancy most in the centre of species’ temperature niches. We estimate that the combined pressures have reduced bumblebee occupancy by 61% across sampled natural habitats, and 65% across human land uses. Our results can inform efforts to conserve bumblebee biodiversity in the face of ongoing land-use changes and accelerating climatic changes.

\newpage

# Introduction

Insect biodiversity has undergone rapid changes in recent decades, although reported trends differ in direction and magnitude. Several studies have suggested steep declines in the abundance, richness and distributions of terrestrial insect species, including bumblebees [@Kerr2015; @Lister2018; @Soroye2020; @vanKlink2020; @Janousek2023]. Other studies have suggested a more mixed picture, with little overall change in insect biodiversity on average [@Outhwaite2020; @Macgregor2019]. Biodiversity changes, including those of insects, are often characterized by strong turnover, with certain species - or groups of species - showing more positive trends than others [@Outhwaite2020; @Kammerer2021; @Halsch2021; @Wagner2021].

Human land use and land-use intensification have been identified as key drivers of insect biodiversity changes. Previous studies have often demonstrated a negative effect on bees (including bumblebees) when natural habitats are converted to agriculture and other human uses [@DePalma2017; @Evans2018; @Samuelson2018; @Janousek2023], although responses vary strongly among species [@Prestele2021; @Cariveau2013] and among regions [@DePalma2016]. Furthermore, within agricultural areas, intensification of farming practices (e.g. lowering of crop diversity, or the removal of flower-rich field margins) is associated with further reductions, on average, in bee diversity [@Hemberger2021; @Kennedy2013]. One facet of agricultural intensification that is of particular importance for bees is the changes in the application of chemical pesticides, often associated with increased toxicity to bees [@Douglas2020], which has been associated with reductions in bee distributions, population persistence, and reproductive performance [@Whitehorn2012; @Woodcock2016; @Siviter2021; @Stuligross2021; @Nicholson2023], although effects vary spatially [@Woodcock2017]. 

Effects of habitat modifications can operate on larger spatial and temporal scales than reflected at the location and time of biodiversity sampling. The availability of natural or semi-natural habitats in the landscape is important for maintaining bee fitness and diversity, including in agricultural habitats [@Evans2018; @Mola2021; @Rutschmann2022; @Proesmans2019; @GutierrezChacon2018]. However, effects often depend on the species and the type of natural habitat in question (e.g., forest or grassland) [@Kammerer2021; @Goulson2010; @GutierrezChacon2018], and are sometimes absent or negative, for example when farmed areas provide abundant floral resources [@Christman2022; @Mola2021]. At the same time, farming practices within agricultural land uses may impact biodiversity within natural habitats. For example, pesticides and other agricultural chemicals may disperse out of the farmland in which they are applied [@vanDerWerf1996]. Similarly, over time, agricultural landscapes tend to become progressively intensified, becoming more homogeneous and losing remnant natural habitats [@Tscharntke2005], effects which have been shown to be associated with a reduction in bumblebee average probability of occurrence in the US Midwest [@Hemberger2021]. Thus, we might expect landscapes that have been farmed for longer to contain a lower diversity of bees and other insects. However, to our knowledge, no previous studies have explicitly tested for a relationship between the length of time landscapes have been substantially modified by human land use and measures of insect biodiversity.

Climate change is already exerting a strong influence on patterns of bee diversity, and its effect is likely to grow rapidly in the coming years. Already, bumblebee distributions have been contracting at the southern range margin [@Kerr2015] and moving to higher elevations [@Pyke2016], while the Community Temperature Index of bumblebee communities (which measures the average temperature affiliation of species in a community) has been increasing [@Fourcade2019], consistent with expectations under climate warming. Indeed, analyses of changes in bee diversity over time have revealed an important role for warming temperatures in observed declines [@Kammerer2021; @Soroye2020; @Janousek2023], and have revealed larger declines in more climatically specialized species [@Williams2009]. Future projections suggest that the impact of climate change on bumblebees is likely to increase in the coming decades [@Prestele2021; @Marshall2018].

Recent evidence points toward synergistic interactive effects of land use, land-use intensification and climate change on biodiversity, but there have been fewer studies focused on insects than on the better-studied vertebrates (e.g., [@Oliver2017; @Hamblin2017; @Fourcade2019; @Waldock2020; @Kammerer2021; @Outhwaite2022; @Janousek2023]). These interactive effects are driven by two key mechanisms. First, habitat disturbance impedes species shifting their ranges in response to climate change [@Oliver2014]. Second, conversion of natural habitats to agriculture and other human uses changes local climates, typically towards hotter and drier conditions with greater extremes of temperature [@Oliver2014; @Williams2020]. For insects in general, the coincidence of intensive agricultural land use and rapid recent climate change is associated with reductions of around 50% of total assemblage abundance, with the greatest losses occurring where little natural habitat remains in the landscape [@Outhwaite2022]. Insect species tolerant of warmer and drier conditions tend, on average, to be more likely to occur within agricultural and urban land uses than other species [@Oliver2017; @Hamblin2017; @Waldock2020], probably as a result of the altered local climatic conditions in these areas. Studies on vertebrates have revealed that populations living close to the upper realized temperature niche of species respond more negatively to human land uses than populations living elsewhere [@Williams2021].

Here we present a continental-scale study, encompassing Europe and North America, simultaneously considering the interacting effects of climate change, land use and land-use intensification on the average probability of occurrence of bumblebee species. A number of studies have tested for large-scale (national or continental extent) effects of climate and land-use variables on bumblebees or other bees. In particular, [@Fourcade2019] assessed change in bumblebee species richness in Norway as a result of climate and land-cover changes, [@Kammerer2021] test the effect on bee species richness of climate, land-cover and insecticide toxic load in three states in the US, and [@Janousek2023] show how climate change, land cover and pesticide application impact the occupancy of the western bumblebee across the US. In our study, we make three further developments over previous work. First, we make use of a recent characterization of the realized climatic niche of bumblebees [@Soroye2020] to test the effect of climate change dependent upon the position of a population within species' realized temperature niche, allowing us to go beyond our recent work considering general interactive effects of land use and climate change across all insects [@Outhwaite2022]. Second, we consider the modelled history of land-use modification to test for the first time whether bumblebee probability of occurrence is affected by the time since landscapes became dominated by human land uses, in addition to considering contemporary local land use, landscape habitat availability and pesticide toxicity. We hypothesize that bumblebee occurrence probability will be lower in agricultural areas, particularly where land-use intensity is higher, where landscapes have been dominated by human activities for longer, and where less natural habitat remains within the landscape. We further hypothesize that occurrence probability will be most strongly reduced in agricultural land use for populations near the upper limit of species' temperature niche, and especially where there have been recent increases in temperature. Third, we use the models to infer how bumblebee probability of occurrence within sampled habitats (both natural habitats and human-dominated land uses) has been shaped by the pressures currently acting on them, allowing us to test whether bumblebee assemblages sampled within apparently natural habitats are depauperate compared to what would be expected in the absence of human pressures.

# Material and Methods

## Data on Responses to Land Use

<!-- SUMMARY STATISTICS: data summaries -->
```{r summarize_data,echo=FALSE,warning=FALSE,results=FALSE}
# Set input directories
dataDir <- "2_PrepareDiversityData/"
modelsDir <- "3_RunModels/"

# Load the compiled species-level data
diversity <- readRDS(paste0(dataDir,"diversity_data.rds"))

allData <- readRDS(paste0(modelsDir,"ModelData.Rds"))

sites <- readRDS(paste0(modelsDir,"AnalysisSites.rds"))

n.studies <- format(length(unique(allData$SS)),big.mark=",",scientific=FALSE)
n.records <- format(nrow(allData),big.mark=",",scientific=FALSE)
n.species <- format(length(unique(allData$Best_guess_binomial)),big.mark=",",
                    scientific=FALSE)
n.countries <- format(length(unique(allData$Country)),big.mark=",",
                      scientific=FALSE)

countries <- unique(allData$Country)

# Put countries in more logical order
countries2 <- na.omit(countries[c(5,1,6,4,10,7,11,13,8,12,9,2,3)])

# Create text string
countries2 <- paste(countries2,collapse=", ")

# Replace last comma with 'and'
countries2 <- sub(",([^,]*)$", " and\\1", countries2)

n.sites <- format(nrow(sites),big.mark=",",scientific=FALSE)

# Get sample years
diversity$SampleYearMid <- as.integer(
  sapply(X = paste(diversity$Sample_midpoint),
         FUN = function(x) strsplit(x,'-')[[1]][1]))

first.year <- min(diversity$SampleYearMid)
last.year <- max(diversity$SampleYearMid)
```

Data describing the occurrence of bumblebees in different land-use types were derived from the PREDICTS database [@Hudson2017], which collates biodiversity records from individual studies that conducted snapshot spatial samples of biodiversity along land-use or land-use-intensity gradients [@Hudson2014]. Studies included in the database had to meet four main criteria: 1) that the sampling methods were published; 2) that biodiversity was sampled at more than one location; 3) that the sampled locations spanned a gradient of land use and/or land-use intensity; and 4) that the sampling protocol was the same across sampled sites. Any records without geographical coordinates were omitted from our analysis. In this study, we focused on the occurrence (presence or absence) of taxa that were identified to the species level.  

The final dataset that we used in this study was derived from `r n.studies` original studies (see Supplementary Material for a list), and consist of `r n.records` records for `r n.species` bumblebee species, collected across `r n.sites` sites in `r n.countries` different countries in North America, and Western, Northern and Southern Europe (Figure S1). The original samples of bumblebees that were collated in the PREDICTS database were collected in the field between `r first.year` and `r last.year`.

## Explanatory Variables

We considered land use in our analysis by dividing sampled sites into either natural or human-modified habitats. In the PREDICTS database, land use is classified into one of six broad categories: primary vegetation (natural habitat with no record of being destroyed), secondary vegetation (habitat recovering to its natural state after past destruction by human actions or extreme natural events), plantation forests (areas used to cultivate woody crops), cropland (areas used to cultivate herbaceous crops, including as fodder for livestock), pasture (areas used regularly or permanently to graze livestock), and urban (areas of human settlement or other buildings, or areas managed for human recreation). We treated primary and secondary vegetation as natural habitats, and all other land-use types as human-modified. Although flying insects such as bumblebees may respond more strongly to landscape habitat availability (variable described below) than to the local land-use conditions [@Janousek2023], local land-use conditions can be important [@Evans2018], and are considered here also because such local land-use differences are fundamental to the sampling structure within the PREDICTS database [@Hudson2014].

We investigated how variables describing surrounding habitat conditions, land-use intensity (specifically, cumulative pesticide toxicity) and species' temperature niche properties shape the occurrence of bumblebees in both natural and human-modified habitats.

Information on habitat condition in the landscapes surrounding sampled sites comprised estimates of the availability of natural habitat, as well as the length of time that landscape habitat has been substantially modified by humans. Availability of surrounding natural habitats was measured as the percentage of natural habitat in the 2 &times; 2-km grid cell within which the biodiversity sample was taken (we tested the robustness of our results when using natural habitat estimated also in 1-km and 5-km grid cells). Original mapped estimates of different land-use types were from a down-scaled land-use projection for 2005 (the mid-point of original field sampling of data included in this analysis; see previous sub-section) at 30-arc-second spatial resolution (approximately 1 km at the Equator) [@Hoskins2016]. These land-use maps estimate the fraction of each grid cell covered by five out of the six land-use types recognised in the PREDICTS database (not plantation forests), and in fact were evaluated against the land-use classifications in the PREDICTS database [@Hoskins2016]. Following our classification of overall land-use type (see above), we summed the primary and secondary vegetation maps to obtain estimates of natural habitat. We projected and resampled this map of the fractional cover of natural habitat at 30-arc-second resolution to an equal-area grid (Behrmann projection) at 1-km resolution, using bilinear interpolation (implemented using the project function of the terra R package Version 1.7-71; [@Hijmans2024]). We then aggregated the resulting 1-km map by factors of two or five, to obtain the maps at 2-km and 5-km grain, calculating the mean fractional cover of natural habitat in the coarser grid cells. Although there is a tendency for landscape natural habitat to be lower for human-modified local land uses compared to natural habitats, this association is weak (ANOVA R^2^ = 0.06 for the natural habitat map at 2-km grain). We derived estimates of the duration of substantial human habitat modification from the land-use harmonisation project land-use reconstruction at 0.5° spatial resolution [@Hurtt2011]. The historical land-use estimates describe reconstructed changes in the fractional cover of primary vegetation, secondary vegetation, croplands, pastures and urban areas from 1500 to 2005. We measured the duration of substantial human habitat modification as the number of years since a 0.5° grid cell was first 30% converted to human-modified land-use types (croplands, pastures or urban areas). Above 30% loss of natural habitats from landscapes, the most sensitive species are typically substantially negatively impacted [@Andren1994]. Grid cells that have never yet reached the conversion threshold of 30% are considered to have a duration of substantial landscape modification of 0 years. However, the appropriate threshold for insect species (including bumblebees) is much less clear than for the better-studied vertebrates. Therefore, we tested the sensitivity of using conversion thresholds of 10% or 50% instead.

To calculate pesticide toxicity, we used modelled estimates of the density (kg/ha) of application of the 20 dominant pesticide active ingredients for each of 6 individual crops (corn, soyabean, wheat, cotton, rice, and alfalfa) or 4 crop groups (vegetables and fruit, orchards and grapes, pasture and hay, and other crops) &ndash; a total of 95 different active ingredients across all crops [@Maggi2019]. These data are presented on a 5-arc-minutes (approximately 10 km at the Equator) global spatial grid. We used the low rather than high modelled estimates from this dataset, to be conservative, but tested the robustness of our models to using the high estimates instead. Estimates of application density for each pesticide active ingredient were divided by estimates of LD50 for that chemical, derived from the Pesticide Properties Database [@Lewis2016], using estimates for honeybees, since this database doesn't record toxicity specifically for bumblebees. We used estimates of contact LD50s where available (87/95 active ingredients), and oral LD50s otherwise (4 active ingredients). For 4 out of 95 pesticide active ingredients where neither a contact nor an oral LD50 was available for honeybees, we excluded that pesticide from the calculations. We then summed the toxicity estimates across all dominant active ingredients for each of the 10 crops/crop groups (i.e., across 200 toxicity maps in total).

For information on species temperature niche properties, we used previously published methods and associated records of bumblebee occurrence [@Soroye2020]. Specifically, we estimated: 1) the exposure of a species' population to extreme high temperatures relative to its realized temperature niche in a baseline period before the onset of rapid climate changes (1901 &mdash; 1975); and 2) the change in exposure caused by climate change between the baseline period and a more recent period (2000 &mdash; 2014). Realized temperature niche limits were estimated for each species as the mean of the five lowest minimum monthly temperatures and the maximum of the five highest maximum monthly temperatures across all bumblebee spatial records in the baseline period. To measure the exposure of each species' population to extreme high temperatures in either the baseline or recent period, we then derived the maximum monthly temperatures for all twelve months in each baseline or recent year. These maximum monthly temperatures were rescaled such that a value of 0 equates to the minimum realized temperature niche limit of the species, and a value of 1 to the maximum niche limit. We then averaged the 12 monthly values to derive an annual mean exposure. Finally, we averaged the annual exposure estimates across all years in either the baseline or recent period. Thus, we were able to estimate, for any biodiversity record, the average exposure to extreme high temperatures in the baseline period, or the change in exposure between the baseline and recent periods. Monthly minimum and maximum temperature estimates for the period 1901 to 2015 were obtained from the CRU global gridded climate reconstruction Version 3.24.01 [@CRU32401], which was the most recent version of this dataset available at the time of calculation.

To control for potential confounding effects of elevation, we also included in the model estimates of elevation at 30-arc-second spatial resolution, obtained from WorldClim Version 2.1 [@Fick2017].

## Statistical Analysis

To analyse the presence or absence of bumblebee species at the sampled locations, we used a binomial (Bernoulli family) Bayesian hierarchical model, accounting for the hierarchical structure of the PREDICTS database. Species were considered absent if they were not recorded but had been targeted by a particular study (typically these were species sampled in at least one other sampled site within a study). As random effects, we fitted: 1) Study identity, to control for differences in sampling protocols and broad geographic differences; 2) Site identity nested within study identity, to control for site-specific factors; and 3) Species identity, to account for variation in responses among species. We were unable to account for variations in sampling effort more explicitly than by incorporating the random effect of study identity, because even for bumblebees the data in the PREDICTS database were collected using many different sampling methods, with non-comparable measures of sampling effort. Although there may be important differences between the two continents in average occurrence probability of bumblebees, as well as in responses to the considered explanatory variables, the dataset was too small to include such variation in the models. As fixed effects, we considered local land use, lanscape-level natural habitat availability, the length of time that the landscape has been substantially modified by human land use, cumulative pesticide toxicity, baseline position with species' thermal niche, and change in this position between the baseline and recent periods. To avoid model over-fitting, given the small size of the dataset, we fitted linear effects for all continuous variables, but a quadratic effect of baseline thermal niche position, because there is strong reason to expect a non-linear effect for this variable. We further considered interactions between local land use and the other continuous variables. We also explored interactions with landscape-level natural habitat availability, but estimated coefficients for these interactions were subject to a high degree of uncertainty, probably owing to slight correlations among the different landscape-level variables. All models were implemented in the brms R package Version 2.21.0 [@Burkner2021]. We ran 4 chains, and assessed convergence by inspection of Rhat values. Chains were run with a burn-in of 1,000 iterations. For comparison, we also built a model containing just land use as a fixed effect. Model explanatory power was assessed using a version of R^2^ values designed for Bayesian models [@Gelman2019]. Collinearity among the explanatory variables included in the models was not high enough to cause major biases in the modelling (Pearson's correlation coefficient, r < 0.55).

To reveal the impacts that environmental changes to date have had on bumblebee occurrence probability across both natural habitats and human-modified habitats, we used our model to infer biodiversity changes across the sampled sites, based on the values of the model explanatory variables at those sites at the time of biodiversity sampling. These estimates thus included the impacts on both natural and modified habitats of climate change, and the modification of landscapes by human land use. Estimates of biodiversity at the time of sampling were expressed as a percentage difference compared to a baseline condition in which we assumed that landscapes were entirely composed of natural habitat, with no history of human modification, subject to no application of pesticides, and with no change in climate. To derive model-estimated differences between baseline and contemporary conditions, with uncertainty, we drew 1,000 sets of coefficient estimates from the posterior samples of the model. Each of these coefficient estimates was used to estimate average bumblebee probability of occurrence under both baseline and contemporary environmental conditions, from which we calculated 1,000 estimates of the percentage difference in average occurrence probability between baseline and contemporary conditions.

# Results

<!-- PLOTTING: Setup for model results plots -->
```{r results_setup, echo=FALSE,warning=FALSE}
#### Model results plot - Setup ####

# set.seed(1234)

# Set input directory
inDir <- "3_RunModels/"

# Load required packages
suppressMessages(suppressWarnings(library(terra)))
suppressMessages(suppressWarnings(library(brms)))
suppressMessages(suppressWarnings(library(Hmisc)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(ggstance)))
suppressMessages(suppressWarnings(library(cowplot)))
suppressMessages(suppressWarnings(library(sf)))
suppressMessages(suppressWarnings(library(snow)))
suppressMessages(suppressWarnings(library(parallel)))

# Log-transform elevation, which is very skewed
diversity$LogElevation <- log(diversity$Elevation+2)

# Log-transform pesticide application, which is also very skewed
diversity$LogPestToxLow <- log(diversity$PestToxLow+1)

diversity$UI <- paste(diversity$LandUse,diversity$Use_intensity,sep="_")
diversity$UI[grepl("Natural",diversity$UI)] <- "Natural"
diversity$UI[grepl("NA",diversity$UI)] <- NA
diversity$UI[grepl("Cannot decide",diversity$UI)] <- NA
diversity$UI <- factor(diversity$UI)
diversity$UI <- relevel(diversity$UI,ref="Natural")

# Select relevant columns
modelData <- diversity[,c('occur','LandUse','TEI_BL',
                          'TEI_delta','LogElevation','SS','SSBS',
                          'Taxon_name_entered','Longitude','Latitude',
                          'Best_guess_binomial','Country',
                          'LogPestToxLow',
                          'NaturalHabitat2k','AgeConv')]

# Remove rows with any NA values
modelData <- na.omit(modelData)

# Load the final model
finalModel <- readRDS(paste0(inDir,"FinalModelBRMS.rds"))
```

<!-- STATISTICS: Model statistics -->
```{r calc_stats,echo=FALSE,warning=FALSE}

# Load the final model with just land use among the main
# explanatory variables
finalModelLU <- readRDS(paste0(inDir,"FinalModelLandUseOnly.rds"))

# Get Bayesian pseudo-R2 values for the final model
r2.marginal <- round(bayes_R2(finalModel,re.form=NA)[,'Estimate'],2)
r2.conditional <- round(bayes_R2(finalModel)[,'Estimate'],2)

# Get the pseudo-r-squared values of the land-use-only model
r2.lu.mod.cond <- round(bayes_R2(finalModelLU)[,'Estimate'],2)
r2.lu.mod.marg <- round(bayes_R2(finalModelLU,re.form=NA)
                        [,'Estimate'],3)

# Get the maximum Rhat value for the estimated coefficients
max.rhat <- max(summary(finalModel)$fixed$Rhat)
```

Availability of natural habitat in the landscape, cumulative pesticide toxicity, land-use history, climatic niche properties and climate change all made a substantial contribution to explaining bumblebee occurrence probability, with effects strongly dependent on local land use (Figure \@ref(fig:figforest)). A model containing just local land use explained only a very small proportion of the variation in bumblebee occurrence (R^2^~marginal~ = `r r2.lu.mod.marg`; R^2^~conditional~ = `r r2.lu.mod.cond`). The full model, containing all of the candidate fixed effects, explained substantially more of the variation in bumblebee occurrence (R^2^~marginal~ = `r r2.marginal`; R^2^~conditional~ = `r r2.conditional`). In the full model, local land use had a negliglble effect on bumblebee probability of occurrence, but crucially played an important role in interaction with the other variables.

<!-- PLOTTING: Forest plot -->
```{r figforest, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from the full binomial Bayesian hierarchical model relating bumblebee occurrence to land use, landscape habitat, pesticide toxicity and thermal niche properties**"}
draws <- as_draws_df(finalModel)[,3:15]

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("2k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxLow1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv1","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

plotCols <- c("#D55E00","#D55E00",
              "#D55E00","#D55E00","#D55E00","#D55E00",
              "#F0E442","#F0E442","#E69F00","#E69F00",
              "#009E73","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Variable,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Variable)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Variable,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Variable),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```

The presence or absence of bumblebee species in both natural habitats and areas of human-modified land use is shaped strongly by habitat characteristics of the surrounding landscape. The availability of natural habitat in the landscape had a signifcant positive effect on probability of occurrence in human-modified land use, but no clear effect within natural habitats (Figure \@ref(fig:figmodeleffects)a). In both local land-use types, bumblebees are more likely to occur in landscapes that have been substantially converted to human land use for a shorter length of time (Figure \@ref(fig:figmodeleffects)b).

<!-- PLOTTING: Model conditional effects -->
```{r figmodeleffects, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=22/2.54,out.width="100%", dpi=1200, fig.cap="**Conditional modelled effects of landscape natural habitat (a), duration of landscape modification (b), pesticide toxicity (c), baseline thermal niche position (d) and change in thermal niche position from climate change (e) on the probability of occurrence of bumblebee species**"}

#### Plotting (natural habitat) ####

# Create data frame with variable values for which to predict occurrence probability
# Set elevation and landscape age variables at their median
# values (to reflect a 'typical' landscape)
# Ignore pesticide toxicity for now (i.e., set to zero)
# Set baseline TEI to the median value
# Set delta TEI to zero (i.e. ignore climate change effects)
nd <- data.frame(LogPestToxLow=0,
                 AgeConv=median(modelData$AgeConv),
                 LogElevation=median(modelData$LogElevation),
                 TEI_BL=median(modelData$TEI_BL),
                 TEI_delta=0)

# Get predicted values with both 67% (~ 1 SE) and 95% credible intervals
preds.67 <- conditional_effects(x = finalModel,effects = "NaturalHabitat2k:LandUse",
                                resolution = 1000,
                                conditions = nd,plot = FALSE,prob=0.67)
preds.95 <- conditional_effects(x = finalModel,effects = "NaturalHabitat2k:LandUse",
                                resolution = 1000,
                                conditions = nd,plot = FALSE,prob=0.95)

plotHab <- ggplot() + 
  geom_ribbon(data = preds.95$`NaturalHabitat2k:LandUse`, 
              mapping = aes(x = NaturalHabitat2k/10,ymin = lower__,
                            ymax = upper__,fill = LandUse),alpha = 0.2) + 
  geom_ribbon(data = preds.67$`NaturalHabitat2k:LandUse`,
              mapping = aes(x = NaturalHabitat2k/10,ymin = lower__,
                            ymax = upper__,fill = LandUse),alpha = 0.4) + 
  geom_line(data = preds.67$`NaturalHabitat2k:LandUse`,
            mapping = aes(x = NaturalHabitat2k/10,y = estimate__,col = LandUse),
            linewidth = 1) + 
  scale_fill_manual(values = c("#009F81","#A40122")) + 
  scale_color_manual(values = c("#009F81","#A40122")) + 
  scale_x_continuous(name = "Natural habitat (%)",limits = c(0,100)) + 
  scale_y_continuous(name = "Probability of occurrence",limits = c(0,1)) + 
  theme_classic() + 
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

#### Plotting (landscape conversion age) ####

# Create data frame with variable values for which to predict occurrence probability
# Set elevation and natural habitat variables at their median
# values (to reflect a 'typical' landscape)
# Ignore pesticide toxicity for now (i.e., set to zero)
# Set baseline TEI to the median value
# Set delta TEI to zero (i.e. ignore climate change effects)
nd <- data.frame(LogPestToxLow=0,
                 NaturalHabitat2k=median(modelData$NaturalHabitat2k),
                 LogElevation=median(modelData$LogElevation),
                 TEI_BL=median(modelData$TEI_BL),
                 TEI_delta=0)

# Get predicted values with both 67% (~ 1 SE) and 95% credible intervals
preds.67 <- conditional_effects(x = finalModel,effects = "AgeConv:LandUse",
                                resolution = 1000,
                                conditions = nd,plot = FALSE,prob=0.67)
preds.95 <- conditional_effects(x = finalModel,effects = "AgeConv:LandUse",
                                resolution = 1000,
                                conditions = nd,plot = FALSE,prob=0.95)

plotAge <- ggplot() + 
  geom_ribbon(data = preds.95$`AgeConv:LandUse`, 
              mapping = aes(x = AgeConv,ymin = lower__,
                            ymax = upper__,fill = LandUse),alpha = 0.2) + 
  geom_ribbon(data = preds.67$`AgeConv:LandUse`,
              mapping = aes(x = AgeConv,ymin = lower__,
                            ymax = upper__,fill = LandUse),alpha = 0.4) + 
  geom_line(data = preds.67$`AgeConv:LandUse`,
            mapping = aes(x = AgeConv,y = estimate__,col = LandUse),
            linewidth = 1) + 
  scale_fill_manual(values = c("#009F81","#A40122")) + 
  scale_color_manual(values = c("#009F81","#A40122")) + 
  scale_x_continuous(name = "Duration of landscape conversion (yrs)") + 
  scale_y_continuous(name = "Probability of occurrence",limits = c(0,1)) + 
  theme_classic() + 
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))


#### Plotting (pesticide toxicity) ####

# Create data frame with variable values for which to predict occurrence probability
# Set elevation, natural habitat and landscape age variables at their median
# values (to reflect a 'typical' landscape)
# Set baseline TEI to the median value
# Set delta TEI to zero (i.e. ignore climate change effects)
nd <- data.frame(NaturalHabitat2k=median(modelData$NaturalHabitat2k),
                 AgeConv=median(modelData$AgeConv),
                 LogElevation=median(modelData$LogElevation),
                 TEI_BL=median(modelData$TEI_BL),
                 TEI_delta=0)

# Get predicted values with both 67% (~ 1 SE) and 95% credible intervals
preds.67 <- conditional_effects(x = finalModel,effects = "LogPestToxLow:LandUse",
                                resolution = 1000,
                                conditions = nd,plot = FALSE,prob=0.67)
preds.95 <- conditional_effects(x = finalModel,effects = "LogPestToxLow:LandUse",
                                resolution = 1000,
                                conditions = nd,plot = FALSE,prob=0.95)

plotPest <- ggplot() + 
  geom_ribbon(data = preds.95$`LogPestToxLow:LandUse`, 
              mapping = aes(x = exp(LogPestToxLow)-1,ymin = lower__,
                            ymax = upper__,fill = LandUse),alpha = 0.2) + 
  geom_ribbon(data = preds.67$`LogPestToxLow:LandUse`,
              mapping = aes(x = exp(LogPestToxLow)-1,ymin = lower__,
                            ymax = upper__,fill = LandUse),alpha = 0.4) + 
  geom_line(data = preds.67$`LogPestToxLow:LandUse`,
            mapping = aes(x = exp(LogPestToxLow)-1,y = estimate__,col = LandUse),
            linewidth = 1) + 
  scale_fill_manual(values = c("#009F81","#A40122")) + 
  scale_color_manual(values = c("#009F81","#A40122")) + 
  scale_x_continuous(name = "Pesticide toxicity") + 
  scale_y_continuous(name = "Probability of occurrence",limits = c(0,1)) + 
  theme_classic() + 
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))


#### Plotting (Baseline TEI) ####

# Create data frame with variable values for which to predict occurrence probability
# Set elevation, natural habitat, and landscape age variables at their median
# values (to reflect a 'typical' landscape)
# Ignore pesticide toxicity for now (i.e., set to zero)
# Set delta TEI to zero (i.e. no climate change)
nd <- data.frame(NaturalHabitat2k=median(modelData$NaturalHabitat2k),
                 LogPestToxLow=0,
                 AgeConv=median(modelData$AgeConv),
                 LogElevation=median(modelData$LogElevation),
                 TEI_delta=0)

# Get predicted values with both 67% (~ 1 SE) and 95% credible intervals
preds.67 <- conditional_effects(x = finalModel,effects = "TEI_BL:LandUse",
                                resolution = 1000,
                                conditions = nd,plot = FALSE,prob=0.67)
preds.95 <- conditional_effects(x = finalModel,effects = "TEI_BL:LandUse",
                                resolution = 1000,
                                conditions = nd,plot = FALSE,prob=0.95)

plotTEI_bl <- ggplot() + 
  geom_ribbon(data = preds.95$`TEI_BL:LandUse`, 
              mapping = aes(x = TEI_BL,ymin = lower__,
                            ymax = upper__,fill = LandUse),alpha = 0.2) + 
  geom_ribbon(data = preds.67$`TEI_BL:LandUse`,
              mapping = aes(x = TEI_BL,ymin = lower__,
                            ymax = upper__,fill = LandUse),alpha = 0.4) + 
  geom_line(data = preds.67$`TEI_BL:LandUse`,
            mapping = aes(x = TEI_BL,y = estimate__,col = LandUse),
            linewidth = 1) + 
  scale_fill_manual(values = c("#009F81","#A40122")) + 
  scale_color_manual(values = c("#009F81","#A40122")) + 
  scale_x_continuous(name = "Baseline thermal position") + 
  scale_y_continuous(name = "Probability of occurrence",limits = c(0,1)) + 
  theme_classic() + 
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))


#### Plotting (delta TEI) ####

# Create data frame with variable values for which to predict occurrence probability
# Set elevation, natural habitat, and landscape age variables at their median
# values (to reflect a 'typical' landscape)
# Ignore pesticide toxicity for now (i.e., set to zero)
# Set baseline TEI to the median value
nd <- data.frame(NaturalHabitat2k=median(modelData$NaturalHabitat2k),
                 LogPestToxLow=0,
                 AgeConv=median(modelData$AgeConv),
                 LogElevation=median(modelData$LogElevation),
                 TEI_BL=median(modelData$TEI_BL))

# Get predicted values with both 67% (~ 1 SE) and 95% credible intervals
preds.67 <- conditional_effects(x = finalModel,effects = "TEI_delta:LandUse",
                                resolution = 1000,
                                conditions = nd,plot = FALSE,prob=0.67)
preds.95 <- conditional_effects(x = finalModel,effects = "TEI_delta:LandUse",
                                resolution = 1000,
                                conditions = nd,plot = FALSE,prob=0.95)

plotTEI_delta <- ggplot() + 
  geom_ribbon(data = preds.95$`TEI_delta:LandUse`, 
              mapping = aes(x = TEI_delta,ymin = lower__,
                            ymax = upper__,fill = LandUse),alpha = 0.2) + 
  geom_ribbon(data = preds.67$`TEI_delta:LandUse`,
              mapping = aes(x = TEI_delta,ymin = lower__,
                            ymax = upper__,fill = LandUse),alpha = 0.4) + 
  geom_line(data = preds.67$`TEI_delta:LandUse`,
            mapping = aes(x = TEI_delta,y = estimate__,col = LandUse),
            linewidth = 1) + 
  scale_fill_manual(values = c("#009F81","#A40122")) + 
  scale_color_manual(values = c("#009F81","#A40122")) + 
  scale_x_continuous(name = "\u0394 thermal position") + 
  scale_y_continuous(name = "Probability of occurrence",limits = c(0,1)) + 
  theme_classic() + 
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

plots <- plot_grid(plotHab,plotAge,plotPest,plotTEI_bl,plotTEI_delta,
                   nrow = 3,ncol = 2,
                   labels =c("a)","b)","c)","d)","e)"),
                   label_x = -0.02,label_size = 12,label_fontface = "plain")

plots

```

Bumblebee occurrence probability is also influenced by the toxicity of pesticides applied in the landscape. In natural habitats, the probability of occurrence of bumblebees decreases strongly with pesticide toxicity, whereas in human-modified land uses there is a negligble effect (Figure \@ref(fig:figmodeleffects)c). Modelled effects of pesticide application rate were very similar when using either the low or high estimates of rates from the original dataset [@Maggi2019] (results not shown).

Finally, bumblebee probability of occurrence is strongly influenced by the climatic niche properties of species, and the effects on these of recent climate change. Populations near the centre of the species' temperature  niche showed the greatest declines in probability of occurrence between natural and human-modified habitats, whereas for populations nearer the upper limit of the temperature niche, there was little difference in occurrence probability (Figure \@ref(fig:figmodeleffects)d). An increase in realized temperature niche position caused by climate warming had a weak negative effect on probability of occurrence in natural habitats (Figure \@ref(fig:figmodeleffects)e). In contrast, the effect of climate warming in human land uses was positive (Figure \@ref(fig:figmodeleffects)e).

The individual model chains converged well (all Rhat &le; `r max.rhat`; Figure SX). There was a good correspondence between the model posterior predicted values and observed records of bumblebee presence and absence (Figure SX).

```{r preds_sites, echo=FALSE,results=FALSE}

#### Predictions for sampled locations ####

# Create baseline data-frame with all natural land use, 100% surrounding natural
# habitat (note that 100% is stored as 1,000), no pesticide and no climate change
nd.bl <- modelData
nd.bl$LandUse <- factor("Natural",levels=c("Natural","Human"))
nd.bl$NaturalHabitat2k <- 1000
nd.bl$LogPestToxLow <- 0
nd.bl$AgeConv <- 0
nd.bl$TEI_delta <- 0

# Extract observed data-frame
nd.obs <- modelData

# cl <- makeCluster(detectCores()-1)
cl <- makeCluster(2)
clusterExport(cl = cl,varlist = c("finalModel","nd.bl","nd.obs"))

# Make 1,000 predictions of % difference between observed
# and baseline conditions
preds <- parSapply(cl = cl,X = 1:10,FUN = function(i){
  
  # Draw one of the posterior parameter estimates at random
  i <- sample(x = 1:4000,size = 1,replace = FALSE)
  
  # Use this set of posterior parameter estimates to get predicted
  # probability of occurrence under baseline and observed environmental
  # conditions
  y.bl <- brms::posterior_epred(
    object = finalModel,newdata = nd.bl,re_formula = NA,
    draw_ids = i)[1,]
  
  y.obs <- brms::posterior_epred(
    object = finalModel,newdata = nd.obs,re_formula = NA,
    draw_ids = i)[1,]
  
  y <- ((y.obs/y.bl)*100)-100
  
  return(y)
})

stopCluster(cl)

# Calculate median and credible intervals across replicate predictions
y.median <- apply(X = preds,MARGIN = 1,FUN = median)
y.lower.95 <- apply(X = preds,MARGIN = 1,FUN = quantile,probs=0.025)
y.upper.95 <- apply(X = preds,MARGIN = 1,FUN = quantile,probs=0.975)
y.lower.67 <- apply(X = preds,MARGIN = 1,FUN = quantile,probs=0.16667)
y.upper.67 <- apply(X = preds,MARGIN = 1,FUN = quantile,probs=0.83333)

# Summarize median, and credible intervals of predictions
# across the two land-use types
lu <- factor(c("Natural","Human"),levels=c("Natural","Human"))
lu.means.median <- tapply(X = y.median,INDEX = nd.obs$LandUse,FUN = median)
lu.means.lower.95 <- tapply(X = y.lower.95,INDEX = nd.obs$LandUse,FUN = median)
lu.means.upper.95 <- tapply(X = y.upper.95,INDEX = nd.obs$LandUse,FUN = median)
lu.means.lower.67 <- tapply(X = y.lower.67,INDEX = nd.obs$LandUse,FUN = median)
lu.means.upper.67 <- tapply(X = y.upper.67,INDEX = nd.obs$LandUse,FUN = median)

```

Using the models to infer bumblebee probability of occurrence at the sampled sites, compared to the probability under hypothetical baseline conditions (100% landscape natural habitat, no history of human land use, no application of agricultural chemicals, and no climate change), we estimate that bumblebee probability of occurrence has been reduced by `r -round(lu.means.median['Natural'],0)`% (95% CI: `r -round(lu.means.upper.95['Natural'],0)`% &ndash; `r -round(lu.means.lower.95['Natural'],0)`%) across sampled natural habitats, and by `r -round(lu.means.median['Human'],0)`% (95% CI: `r -round(lu.means.upper.95['Human'],0)`% &ndash; `r -round(lu.means.lower.95['Human'],0)`%) across sampled human land uses (Figure \@ref(fig:figpredssites)).

```{r figpredssites, echo=FALSE, warning=FALSE, fig.width=12.5/2.54, fig.height=13/2.54,out.width="60%", dpi=1200, fig.cap="**Model-inferred differences in average bumblebee probability of occurrence at sampled sites in natural (primary and secondary) habitats and human-modified (agricultural and urban) habitats, compared to hypothetical baseline conditions (natural land use across the whole landscape, no human land-use history, no agricultural chemical application, and no climate change)**. Predictions were generated for every species recorded at each sampled site across both natural and human-modified habitats, based on the values of the explanatory variables as estimated for the sampled sites. Points show the median prediction, while thick and thin error bars, respectively, represent the 67% and 95% confidence intervals. Predictions capture both model and parameter uncertainty, by drawing 1000 sets of estimates both across models (weighted by AIC weight), and within sampled models according to the modelled uncertainty in the parameter estimates"}

# Average predicted differences in occurrence probability across sampled sites
y.median.sites <- tapply(X = y.median,INDEX = nd.obs$SSBS,FUN = mean)
y.upper.sites <- tapply(X = y.upper.95,INDEX = nd.obs$SSBS,FUN = mean)
y.lower.sites <- tapply(X = y.lower.95,INDEX = nd.obs$SSBS,FUN = mean)

# Create data frame for site-level averages
sites <- data.frame(y=y.median.sites,yplus=y.upper.sites,yminus=y.lower.sites,
                    SSBS=names(y.median.sites))
sites$Longitude <- nd.obs$Longitude[match(sites$SSBS,nd.obs$SSBS)]
sites$Latitude <- nd.obs$Latitude[match(sites$SSBS,nd.obs$SSBS)]

# Create basemap for North American and Western European sub-regions
baseMap <- vect("0_data/UN_subregion.shp")
baseMap <- terra::subset(baseMap,baseMap$SUBREGION %in% c(21,13,154,155,39))
baseMap <- st_as_sf(baseMap)

# Create map of predictions for sites sampled in PREDICTS
predsMap <- vect(x = sites,geom=c("Longitude","Latitude"),
                 crs = '+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0')
predsMap <- st_as_sf(predsMap)

# Plot base map and predictions for sampled locations
mapPreds <- ggplot() + geom_sf(data = baseMap,fill = "#ffffff") + 
  geom_sf(data=predsMap,mapping = aes(colour=y)) + 
  scale_color_continuous("\u0394 Probability of\noccurrence (%)",type = "viridis") + 
  scale_x_continuous(limits = c(-180,60)) + 
  theme_classic() + 
  theme(axis.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.text = element_text(size=10))

# Create error-bar plot showing predictions at sampled locations
# by land-use type
plotLU <- ggplot() + 
  geom_linerange(mapping = aes(x = lu,
                               ymin = lu.means.lower.95,
                               ymax = lu.means.upper.95,
                               col = lu)) +
  geom_pointrange(mapping = aes(x = lu,
                                y = lu.means.median,
                                ymin = lu.means.lower.67,
                                ymax = lu.means.upper.67,
                                col = lu),
                  linewidth = 1, size = 0.8) +
  scale_colour_manual(values = c("#009F81","#A40122")) +
  scale_y_continuous(name = "\u0394 Probability of occurrence (%)",
                     limits = c(-100,0)) + 
  scale_x_discrete(name = "Land use") + 
  geom_hline(mapping = aes(yintercept = 0),
             alpha = 0.3, linetype = 'dashed') +
  theme_classic() + 
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 12))

# Create final figure
plots <- cowplot::plot_grid(mapPreds,plotLU,nrow = 2,ncol = 1,
                   rel_heights = c(0.4,0.6),rel_widths = c(1,0.1))
plots
```

# Discussion

Our results reveal important interactive effects of climate change and land use on bumblebee biodiversity. There is increasing recognition that synergistic interactions among pressures, especially land-use-climate interactions, may be exacerbating ongoing biodiversity changes [@Oliver2014; @Williams2020], but relatively few studies have investigated such interactions in the context of insect biodiversity change [@Oliver2017; @Hamblin2017; @Waldock2020; @Outhwaite2022]. While a correlative empirical analysis such as ours cannot confidently elucidate mechanisms, it is likely that two mechanisms play a key role: microclimatic changes associated with land-use change; and habitat fragmentation impeding responses to climate change [@Oliver2014]. We estimate that bumblebee occurrence probability has been reduced substantially across both natural habitats and human-modified areas. As climate changes intensify, and further modification of land use occurs, changes in bumblebee biodiversity are likely to accelerate.

We find that bumblebee probability of occurrence is reduced most in areas of human land use, especially in landscapes with a lower cover of natural habitats that have been modified by humans for a long period of time. Our findings are consistent with previous, local or regional-scale studies, which generally found negative effects of human land use, and positive effects of landscape semi-natural habitat on bumblebee diversity [@Samuelson2018; @Proesmans2019; @Janousek2023]. By analysing a collation of data spanning two continents, we show the generality of these effects, and also highlight the importance of landscape land-use history in shaping responses. Importantly, our findings show that responses to land use interact strongly with climate change and with the position of populations within species' climatic niches.

Interestingly, bumblebee occurrence probability is reduced most by human land use in the centre of species' temperature niches. Previous research on vertebrates has shown that populations are more likely to be lost in human-disturbed land uses near the edge of species' thermal niches [@Williams2021]. At the same time, a study of insects showed that species affiliated with cooler and wetter conditions were more likely to be absent in human land uses than those affiliated with warmer and drier conditions [@Waldock2020]. These previous results are expected, given that human-disturbed land uses have hotter and drier local climates than natural habitats [@Frishkoff2015; @Senior2017].

Climate warming, which moves populations closer to species' upper thermal limits, reduced occurrence probability in natural habitats, but not consistently in human land uses. This contrasts with the results of a study on birds in North America, where declines in response to habitat loss were strongest where summer temperatures had warmed the most  [@Northrup2019]. The strong negative response of bumblebees to human land use in the temperature niche centre, and the negative effect of climate change in natural habitats, may be associated with factors such as resource limitation [@Requier2020; @McCombs2022], rather than local climatic changes. At the same time, the relative resilience of bumblebee populations to land use at the upper edge of species' thermal niche possibly points towards some local adaptation to climatic conditions, as suggested by some previous studies [@Jackson2020], but see [@Martinet2021]. Overall, the interactive effects of climate change and land-use change on occurrence probability across the niches of bumblebee species appear more complicated than for other groups of species.

Surprisingly, land-use intensification, measured here in terms of pesticide and fertilizer application rates, had relatively little effect on bumblebee biodiversity. However, we caution that the available global data on application of agricultural chemicals are resolved only at a coarse spatial resolution of 5 arc-minutes (approximately 10 km at the Equator) [@Mueller2012; @Maggi2019]. Similar studies to ours have also found weak effects of fertilizer application rate using the same dataset [@Outhwaite2022], while a study of UK bees found detectable, but weak, effects of pesticide (specifically neonicotinoid) exposure on population trends [@Woodcock2016]. The results of these analyses of field data, using coarse-scale estimates of pesticide application rates, contrast strongly with results based on experimental exposure of bees to neonicotinoid pesticides [@Whitehorn2012; @Stuligross2021]. There is thus an urgent need for more detailed and finely resolved continental or global spatial data on the application of specific pesticides.

Our results highlight the growing impacts on biodiversity from interactions among pressures, which will likely intensify rapidly in the coming decades. Future climate warming will likely lead to decreases, on average, in bumblebee diversity [@Prestele2021; @Marshall2018; @Soroye2020], with our results highlighting particularly important declines in natural habitats. While, in general, Europe and North America are expected to see a reduction in cropland area in future, with relatively little change in agricultural intensity compared to other world regions [@Zabel2019], ongoing changes in agricultural practices are highly likely. We highlight the complex interplay of land-use and climatic factors in shaping bumblebee biodiversity, which need to be considered when predicting changes in biodiversity. Our findings also point to some potential interventions that could mitigate losses of bumblebee biodiversity. In particular, efforts to return areas of agriculture to a natural state, and to restore patches of natural habitat within farmed landscapes, are expected to benefit bumblebees, and also other insect groups [@Outhwaite2022]. However, any efforts to improve bumblebee biodiversity through changes in land use are unlikely to be successful unless the rate of climate change is also reduced.

Using our models to infer historical changes in bumblebee probability of occurrence suggests that modification of landscapes, combined with recent climate change, has reduced bumblebee biodiversity within natural habitats almost as much as in human-modified areas. While agricultural areas experience the direct impacts of habitat loss, many natural habitats are exposed to degradation of their surrounding landscapes, while both natural and human-modified areas are exposed to the effects of climate change. Land-use impacts are often quantified using spatial analyses, comparing sampled biodiversity between natural and modified areas [@Newbold2015]. The fact that natural habitats are composed of depauperate communities, heavily impacted by changes in their landscapes, means that such estimates are likely to underestimate land-use impacts substantially [@Newbold2019].

Understanding changes in biodiversity through time using a spatial database of species records is inevitably subject to certain limitations. Most importantly, such an analysis does not permit a consideration of the dynamics of biodiversity change [@DePalma2018]. However, time-series data for insects are lacking in most regions. Another limitation is the reliance of large-scale studies on coarsely resolved estimates of the pressures on biodiversity. While the availability of fine-scale land-use/land-cover data has improved rapidly in recent years [@Jung2020], we still lack fine-scale estimates of the application of agricultural chemicals, a particularly important pressure on flying insect biodiversity [@Wagner2021]. To estimate the position of populations within species' temperature niches, we use information on species' realized distributions with respect to temperature. Such realized temperature limits may not correspond with the fundamental physiological limits of species, but estimates of the latter [@Bennett2018] are available for too few species to be useful in large-scale analyses. A previous study showed that realized thermal limits are clearly associated with occupancy change in response to recent climate change [@Soroye2020], suggesting these limits are ecologically important. Finally, all correlative analyses are unable to prove causal mechanisms. The complexity of the interactive effects we document highlights a need for more experimental studies or detailed field-level assessments to investigate the mechanistic basis for the patterns we report.

In conclusion, we show that interactions between climate change and land use drive rapid and substantial changes in bumblebee biodiversity across both natural habitats and human-modified land. Given the importance of bumblebees for pollinating wild plants [@Ollerton2017] and agricultural crops [@Klein2007], these changes are likely to have important effects on both natural ecosystems and on our ability to grow food. Mitigating further losses of bumblebee diversity, and potentially reversing past declines, will require strong mitigation of climate change combined with restoration of natural habitats. To predict future changes in biodiversity, including of bumblebees, we must account for the complex interaction of climate and land-use pressures that are currently affecting species.

<!--
```{r session_info, echo=FALSE,reulst=TRUE,eval=TRUE}
print(sessionInfo())
```
-->

# Data and Code Availability:

Code to run the analyses is available on GitHub ([https://github.com/timnewbold/BumblebeesLUClimStudy](https://github.com/timnewbold/BumblebeesLUClimStudy)). Data are available on FigShare ([https://figshare.com/s/5e88f431142bc8fc454a](https://figshare.com/s/5e88f431142bc8fc454a)) - temporary link for peer review.

# Acknowledgments

This research was supported by a Royal Society International Exchanges grant to TN and JK (IE161031), UK Natural Environment Research Council grants to TN (NE/R010811/1 and NE/V006533/1) and a Royal Society University Research Fellowship to TN (UF150526). The authors declare no competing interests.

# References
