---
title: Supplementary Material for 'Bumblebee occupancy responds to complex interactions between local and landscape land use, climatic niche properties and climate change'
author: "Tim Newbold^1,\\*^, Jeremy Kerr^2^, Peter Soroye^2,&dagger;^, Jessica J. Williams^1^"
output:
  bookdown::word_document2:
    toc: false
    fig_caption: TRUE
    number_sections: no
always_allow_html: true
csl: science-advances.csl
bibliography: bumblebees_lu_cc_refs.bib
---

^1^Centre for Biodiversity and Environment Research, Department of Genetics, Evolution and Environment, University College London, London, UK

^2^Department of Biology, University of Ottawa, Ottawa, Ontario, Canada

^&dagger;^Current address: Wildlife Conservation Society Canada, Toronto, Ontario, Canada

^\*^Correspondence: [t.newbold@ucl.ac.uk](mailto:t.newbold@ucl.ac.uk)

<!-- SETUP: Global options -->
```{r,echo=FALSE,results=FALSE}
knitr::opts_chunk$set
options(scipen=20)
set.seed(1234)
```

# Model Diagnostic Checks

<br />    

<!-- SETUP: read data and models -->
```{r,echo=FALSE,results=FALSE,warning=FALSE}
# Set input directories
dataDir <- "2_PrepareDiversityData/"
modelsDir <- "3_RunModels/"
bestModelDir <- "6_RefitBestModel/"
mcmcModelDir <- "7_FitMCMCModel/"

# Load the compiled species-level data
diversity <- readRDS(paste0(dataDir,"diversity_data.rds"))

allData <- readRDS(paste0(modelsDir,"ModelData.Rds"))

sites <- readRDS(paste0(modelsDir,"AnalysisSites.rds"))

# Set input directory
inDir <- "3_RunModels/"

# Load required packages
suppressMessages(suppressWarnings(library(StatisticalModels)))

# Log-transform elevation, which is very skewed
diversity$LogElevation <- log(diversity$Elevation+2)

# Log-transform fertilizer application, which is also very skewed
diversity$LogFertilizer <- log(diversity$Fertilizer+1)

diversity$UI <- paste(diversity$LandUse,diversity$Use_intensity,sep="_")
diversity$UI[grepl("Natural",diversity$UI)] <- "Natural"
diversity$UI[grepl("NA",diversity$UI)] <- NA
diversity$UI[grepl("Cannot decide",diversity$UI)] <- NA
diversity$UI <- factor(diversity$UI)
diversity$UI <- relevel(diversity$UI,ref="Natural")

# Select relevant columns
modelData <- diversity[,c('occur','LandUse','TEI_BL',
                          'TEI_delta','LogElevation','SS','SSBS',
                          'Taxon_name_entered','Longitude','Latitude',
                          'Best_guess_binomial','Country',
                          'Pesticide','LogFertilizer',
                          'NaturalHabitat','AgeConv')]
# Remove rows with any NA values
modelData <- na.omit(modelData)

# Load the final model
finalModels <- readRDS(paste(inDir,"FinalModels.rds",sep=""))

bestModel <- readRDS(paste0(bestModelDir,"BestModel.Rds"))
```

<!-- PLOTTING: model diagnostic checks -->
```{r diagnostic,echo=FALSE,results=FALSE,warning=FALSE, fig.width=8.5, fig.height=8,out.width="50%", fig.cap="**Q-Q plot of scaled model residuals compared to the expected distribution of residuals**. This plot was produced directly from the DHARMa R package Version 0.4.6"}
suppressMessages(suppressWarnings(library(DHARMa)))
suppressMessages(suppressWarnings(library(StatisticalModels)))

simResids <- simulateResiduals(finalModels[[17]])

par(cex=2)
par(cex.axis=1.3)
par(cex.lab=1.3)
par(cex.text=1.3)
par(las=1)
par(tck=-0.01)
par(mgp=c(2.2,0.2,0))
par(mar=c(3.4,3.4,0.2,0.2))

DHARMa::plotQQunif(simResids,main=NA)
box(bty="l")

```

<br />

```{r lme4mcmc,echo=FALSE,results=FALSE,warning=FALSE, fig.width=8.5, fig.height=7,out.width="50%", fig.cap="**Comparison of fitted values from the main best-fitting model and a model with identical structure fitted using MCMC**. This comparison was made to check for robustness of model-estimated probabailities of presence, given the violation of the assumptions of standard parametric statistical tests, shown in Figure S1. The MCMC model was fit using the MCMCglmm R package Version 2.34"}
predsCompared <- readRDS(file = paste0(mcmcModelDir,"MCMCLME4PredictionsCompared.Rds"))

par(las=1)
par(tck=-0.01)
par(mgp=c(1.8,0.2,0))
par(mar=c(2.8,2.8,0.2,0.2))
par(cex=2)
par(cex.axis=1.3)
par(cex.lab=1.3)
par(cex.text=1.3)
plot(predsCompared$predsLM4,predsCompared$predsMCMC,pch=16,cex=0.2,xlab="Original Fitted Values",ylab="MCMC Fitted Values")
abline(0,1,col="#ff0000",lwd=2)

```

<br /> 

```{r spatautocorrel,echo=FALSE,results=FALSE,warning=FALSE, fig.width=8.5, fig.height=7,out.width="50%", fig.cap="**Distribution of P values from a series of Moran's tests for spatial autocorrelation in the residuals associated with each individual study**. For each study, we calculated the average residual for each sampled site, and then ran a Moran's test for spatial autocorrelation in these average residuals. Moran's tests were performed using the spdep R package Version 1.2-8"}

bestModel$data <- droplevels(bestModel$data)

siteAvgResids <- data.frame(resids=tapply(X = residuals(bestModel$model),
                                          INDEX = bestModel$data$SSBS,
                                          FUN = mean))
siteAvgResids$SSBS <- row.names(siteAvgResids)
row.names(siteAvgResids) <- NULL
siteAvgResids$Longitude <- bestModel$data$Longitude[
  match(siteAvgResids$SSBS,bestModel$data$SSBS)]
siteAvgResids$Latitude <- bestModel$data$Latitude[
  match(siteAvgResids$SSBS,bestModel$data$SSBS)]
siteAvgResids$SS <- bestModel$data$SS[
  match(siteAvgResids$SSBS,bestModel$data$SSBS)]

moran <- as.data.frame(t(sapply(X = as.character(unique(siteAvgResids$SS)),FUN = function(ss){
  
  data.sub <- droplevels(siteAvgResids[siteAvgResids$SS==ss,])
  
  ds.nb<-try(dnearneigh(cbind(data.sub$Longitude,data.sub$Latitude),
                            d1=0.00000001,d2=10),silent=TRUE)
  ds.listw<-try(nb2listw(ds.nb),silent=TRUE)
  mt<-tryCatch(moran.test(data.sub$resids,ds.listw),silent=TRUE,error=function(e) e,
               warning=function(w) w)
  
  if(class(mt)[1]=="htest"){
        if ((!is.na(mt$statistic))){
          return(data.frame(SS=ss,i=mt$statistic,p=mt$p.value))
        } else {
          return(data.frame(SS=ss,i=NA,p=NA))
        }
    
  } else {
    return(data.frame(SS=ss,i=NA,p=NA))
  }
  
})))

moran$SS <- as.character(moran$SS)
moran$i <- as.numeric(moran$i)
moran$p <- as.numeric(moran$p)

par(las=1)
par(tck=-0.01)
par(mgp=c(1.8,0.2,0))
par(mar=c(2.8,2.8,0.2,0.2))
par(cex=2)
par(cex.axis=1.3)
par(cex.lab=1.3)
par(cex.text=1.3)
hist(x = moran$p,breaks = seq(from=0,to=1,by=0.05),main=NA,xlab="Moran's Test P Value")
box(bty="l")
abline(v=0.05,col="#ff0000",lwd=3)

text(x=0.07,y=8,labels=paste0("Significant \nspatial \nautocorrelation \nin ",round(length(which(moran$p<0.05))/length(which(!is.na(moran$p)))*100,1),"% \nof studies"),
     pos=4)

```