---
title: Supplementary Material for 'Bumblebee probability of occurrence responds to interactions between local and landscape land use, climatic niche properties and climate change'
author: "Tim Newbold^1,\\*^, Jeremy Kerr^2^, Peter Soroye^2,&dagger;^, Jessica J. Williams^1^"
output:
  bookdown::word_document2:
    toc: false
    fig_caption: TRUE
    number_sections: no
always_allow_html: true
csl: ecology-letters.csl
bibliography: bumblebees_lu_cc_refs.bib
---

^1^Centre for Biodiversity and Environment Research, Department of Genetics, Evolution and Environment, University College London, London, UK

^2^Department of Biology, University of Ottawa, Ottawa, Ontario, Canada

^&dagger;^Current address: Wildlife Conservation Society Canada, Toronto, Ontario, Canada

^\*^Correspondence: [t.newbold@ucl.ac.uk](mailto:t.newbold@ucl.ac.uk)

<!-- SETUP: Global options -->
```{r,echo=FALSE,results=FALSE}
knitr::opts_chunk$set
options(scipen=20)
set.seed(1234)
```

<!-- SETUP: read data and models -->
```{r,echo=FALSE,results=FALSE,warning=FALSE}
# Load required packages
suppressMessages(suppressWarnings(library(terra)))
suppressMessages(suppressWarnings(library(brms)))
suppressMessages(suppressWarnings(library(Hmisc)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(ggstance)))
suppressMessages(suppressWarnings(library(cowplot)))
suppressMessages(suppressWarnings(library(sf)))
suppressMessages(suppressWarnings(library(snow)))
suppressMessages(suppressWarnings(library(parallel)))
suppressMessages(suppressWarnings(library(spdep)))

# Set input directories
dataDir <- "2_PrepareDiversityData/"
modelsDir <- "3_RunModels/"
modelsDirRobustness <- "6_ModelRobustnessChecks/"

# Load the compiled species-level data
diversity <- readRDS(paste0(dataDir,"diversity_data.rds"))

allData <- readRDS(paste0(modelsDir,"ModelData.Rds"))

sites <- readRDS(paste0(modelsDir,"AnalysisSites.rds"))

# Set input directory
inDir <- "3_RunModels/"

# Log-transform elevation, which is very skewed
diversity$LogElevation <- log(diversity$Elevation+2)

# Log-transform pesticide application, which is also very skewed
diversity$LogPestToxLow <- log(diversity$PestToxLow+1)

diversity$UI <- paste(diversity$LandUse,diversity$Use_intensity,sep="_")
diversity$UI[grepl("Natural",diversity$UI)] <- "Natural"
diversity$UI[grepl("NA",diversity$UI)] <- NA
diversity$UI[grepl("Cannot decide",diversity$UI)] <- NA
diversity$UI <- factor(diversity$UI)
diversity$UI <- relevel(diversity$UI,ref="Natural")

# Select relevant columns
modelData <- diversity[,c('occur','LandUse','TEI_BL',
                          'TEI_delta','LogElevation','SS','SSBS',
                          'Taxon_name_entered','Longitude','Latitude',
                          'Best_guess_binomial','Country',
                          'LogPestToxLow',
                          'NaturalHabitat2k','AgeConv')]

# Remove rows with any NA values
modelData <- na.omit(modelData)

# Load the final model
finalModel <- readRDS(paste0(inDir,"FinalModelBRMS.rds"))
```

# Sites in Analysis

<br />

<!-- PLOTTING: Plot map of sites included in analysis -->
```{r figsitesmap, echo=FALSE, warning=FALSE, fig.width=16.5/2.54, fig.height=12/2.54,dpi=1200, out.width="100%", fig.cap="**Location of sites used in the analysis**. Base map shows the UN sub-regions from which bumblebee data were obtained: Northern America, Central America, Northern Europe, Western Europe and Southern Europe. Note that Mexico is classed within Central America in the UN sub-region scheme. Map is plotted with geographical coordinates using the WGS 1984 datum"}

dataDir <- "0_data/"

# Create basemap for North American and Western European sub-regions
baseMap <- vect("0_data/UN_subregion.shp")
baseMap <- terra::subset(baseMap,baseMap$SUBREGION %in% c(21,13,154,155,39))
baseMap <- st_as_sf(baseMap)

# Create map of the sites sampled in PREDICTS
sitesMap <- vect(x = sites,geom=c("Longitude","Latitude"),
                 crs = '+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0')

sitesMap <- st_as_sf(sitesMap)

# Plot base map and overlaid with sampled sites
mapPreds <- ggplot() + geom_sf(data = baseMap,fill = "#ffffff") + 
  geom_sf(data=sitesMap,colour = "#cc0000") +
  scale_x_continuous(limits = c(-180,60)) + 
  theme_classic() + 
  theme(axis.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.text = element_text(size=10))

mapPreds
```

\newpage

# Distributions of Explanatory Variables

<br />

```{r explVarsHists,echo=FALSE,warning=FALSE,results=FALSE,fig.width=16.5/2.54,fig.height=21.5/2.54,out.width="100%",dpi=1200,fig.cap="**Distributions of explanatory variables in the modelled data**"}

histNatHab <- ggplot() + geom_histogram(data = modelData,mapping = aes(x = NaturalHabitat2k/10,fill=LandUse),
                                        bins=10,position="dodge") + scale_x_continuous(name = "Natural Habitat (%)") + scale_y_continuous(name = "Frequency") + theme(axis.title = element_text(size = 12),axis.text = element_text(size = 10)) + theme_classic()

histAge <- ggplot() + geom_histogram(data = modelData,mapping = aes(x = AgeConv,fill=LandUse),
                                     bins=10,position="dodge") + scale_x_continuous(name = "Duration of landscape conversion (yrs)") + scale_y_continuous(name = "Frequency") + theme(axis.title = element_text(size = 12),axis.text = element_text(size = 10)) + theme_classic()

histPest <- ggplot() + geom_histogram(data = modelData,mapping = aes(x = exp(LogPestToxLow)-1,fill=LandUse),
                                      bins=10,position="dodge") + scale_x_continuous(name = "Pesticide toxicity") + scale_y_continuous(name = "Frequency") + theme(axis.title = element_text(size = 12),axis.text = element_text(size = 10)) + theme_classic()

histTEIBL <- ggplot() + geom_histogram(data = modelData,mapping = aes(x = TEI_BL,fill=LandUse),
                                       bins=10,position="dodge") + scale_x_continuous(name = "Baseline thermal position") + scale_y_continuous(name = "Frequency") + theme(axis.title = element_text(size = 12),axis.text = element_text(size = 10)) + theme_classic()

histTEIDelta <- ggplot() + geom_histogram(data = modelData,mapping = aes(x = TEI_delta,fill=LandUse),
                                          bins=10,position="dodge") + scale_x_continuous(name = "\u0394 thermal position") + scale_y_continuous(name = "Frequency") + theme(axis.title = element_text(size = 12),axis.text = element_text(size = 10)) + theme_classic()

plots <- plot_grid(histNatHab,histAge,histPest,histTEIBL,histTEIDelta,
                   nrow = 3,ncol = 2)

plots
```

# Model Diagnostic Checks

<br />    

```{r diagnostic setup,echo=FALSE,results=FALSE,warning=FALSE}

vars <- variables(finalModel)

vars <-  vars[grepl("^b_",vars)]

```

<!-- PLOTTING: model diagnostic checks -->
```{r diagnostic1,echo=FALSE,warning=FALSE, fig.width=16.5/2.54, fig.height=22.5/2.54,dpi=1200,out.width="100%"}

plot(finalModel,variable=vars[1:4])

```

```{r diagnostic2,echo=FALSE,warning=FALSE, fig.width=16.5/2.54, fig.height=22.5/2.54,dpi=1200,out.width="100%"}

plot(finalModel,variable=vars[5:8])

```

```{r diagnostic3,echo=FALSE,warning=FALSE, fig.width=16.5/2.54, fig.height=22.5/2.54,dpi=1200,out.width="100%"}

plot(finalModel,variable=vars[9:12])

```

```{r diagn4,echo=FALSE,results=FALSE,warning=FALSE,fig.width=16.5/2.54,fig.height=16.875/2.54,dpi=1200,out.width="100%",fig.cap="**Test**"}

plot(finalModel,variable=vars[13:15])

```
<br />

```{r ppcheck,echo=FALSE,results=FALSE,warning=FALSE,fig.width=8.25/2.54,fig.height=8/2.54,dpi=1200,out.width="50%",fig.cap="**Density plot comparing observed presences/absences with model-predicted values drawn from the posterior distribution**. Figure is produced directly by the *pp_check* function of the brms R package Version 2.21.0"}
pp_check(finalModel,ndraws=100)
```

<br /> 

```{r spatautocorrel,echo=FALSE,results=FALSE,warning=FALSE, fig.width=8.5, fig.height=7,out.width="50%", fig.cap="**Distribution of P values from a series of Moran's tests for spatial autocorrelation in the residuals associated with each individual study**. For each study, we calculated the average residual for each sampled site, and then ran a Moran's test for spatial autocorrelation in these average residuals. Moran's tests were performed using the spdep R package Version 1.2-8"}

modelData <- droplevels(modelData)

siteAvgResids <- data.frame(resids=tapply(X = residuals(finalModel)[,1],
                                          INDEX = modelData$SSBS,
                                          FUN = mean))
siteAvgResids$SSBS <- row.names(siteAvgResids)
row.names(siteAvgResids) <- NULL
siteAvgResids$Longitude <- modelData$Longitude[
  match(siteAvgResids$SSBS,modelData$SSBS)]
siteAvgResids$Latitude <- modelData$Latitude[
  match(siteAvgResids$SSBS,modelData$SSBS)]
siteAvgResids$SS <- modelData$SS[
  match(siteAvgResids$SSBS,modelData$SSBS)]

moran <- as.data.frame(t(sapply(X = as.character(unique(siteAvgResids$SS)),FUN = function(ss){
  
  data.sub <- droplevels(siteAvgResids[siteAvgResids$SS==ss,])
  
  ds.nb<-try(dnearneigh(cbind(data.sub$Longitude,data.sub$Latitude),
                            d1=0.00000001,d2=10),silent=TRUE)
  ds.listw<-try(nb2listw(ds.nb),silent=TRUE)
  mt<-tryCatch(moran.test(data.sub$resids,ds.listw),silent=TRUE,error=function(e) e,
               warning=function(w) w)
  
  if(class(mt)[1]=="htest"){
        if ((!is.na(mt$statistic))){
          return(data.frame(SS=ss,i=mt$statistic,p=mt$p.value))
        } else {
          return(data.frame(SS=ss,i=NA,p=NA))
        }
    
  } else {
    return(data.frame(SS=ss,i=NA,p=NA))
  }
  
})))

moran$SS <- as.character(moran$SS)
moran$i <- as.numeric(moran$i)
moran$p <- as.numeric(moran$p)

par(las=1)
par(tck=-0.01)
par(mgp=c(1.8,0.2,0))
par(mar=c(2.8,2.8,0.2,0.2))
par(cex=2)
par(cex.axis=1.3)
par(cex.lab=1.3)
par(cex.text=1.3)
hist(x = moran$p,breaks = seq(from=0,to=1,by=0.05),main=NA,xlab="Moran's Test P Value")
box(bty="l")
abline(v=0.05,col="#ff0000",lwd=3)

text(x=0.07,y=8,labels=paste0("Significant \nspatial \nautocorrelation \nin ",round(length(which(moran$p<0.05))/length(which(!is.na(moran$p)))*100,1),"% \nof studies"),
     pos=4)

```

<br />

# Model Robustness Checks

<br />

```{r figforestnathab1, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from the full binomial Bayesian hierarchical model relating bumblebee occurrence to land use, landscape habitat, pesticide toxicity and thermal niche properties**. From the model testing robustness to fitting estimates of landscape natural habitat at 1-km instead of 2-km spatial grain"}
finalModel_nathab1 <- readRDS(paste0(modelsDirRobustness,"ModelNatHab1k.rds"))

draws <- as_draws_df(finalModel_nathab1)[,3:15]

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("1k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxLow1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv1","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

plotCols <- c("#D55E00","#D55E00",
              "#D55E00","#D55E00","#D55E00","#D55E00",
              "#F0E442","#F0E442","#E69F00","#E69F00",
              "#009E73","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Variable,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Variable)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Variable,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Variable),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```

```{r figforestnathab5, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from the full binomial Bayesian hierarchical model relating bumblebee occurrence to land use, landscape habitat, pesticide toxicity and thermal niche properties**. From the model testing robustness to fitting estimates of landscape natural habitat at 5-km instead of 2-km spatial grain"}
finalModel_nathab5 <- readRDS(paste0(modelsDirRobustness,"ModelNatHab5k.rds"))

draws <- as_draws_df(finalModel_nathab5)[,3:15]

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("5k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxLow1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv1","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

plotCols <- c("#D55E00","#D55E00",
              "#D55E00","#D55E00","#D55E00","#D55E00",
              "#F0E442","#F0E442","#E69F00","#E69F00",
              "#009E73","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Variable,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Variable)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Variable,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Variable),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```

```{r figforestage10, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from the full binomial Bayesian hierarchical model relating bumblebee occurrence to land use, landscape habitat, pesticide toxicity and thermal niche properties**. From the model testing robustness to fitting estimates of duration of substantial habitat modification where the threshold for defining substantial habitat modification is set at 10% instead of 30%"}
finalModel_age10 <- readRDS(paste0(modelsDirRobustness,"ModelAge10.rds"))

draws <- as_draws_df(finalModel_age10)[,3:15]

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("2k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxLow1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv101","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

plotCols <- c("#D55E00","#D55E00",
              "#D55E00","#D55E00","#D55E00","#D55E00",
              "#F0E442","#F0E442","#E69F00","#E69F00",
              "#009E73","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Variable,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Variable)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Variable,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Variable),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```

```{r figforestage50, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from the full binomial Bayesian hierarchical model relating bumblebee occurrence to land use, landscape habitat, pesticide toxicity and thermal niche properties**. From the model testing robustness to fitting estimates of duration of substantial habitat modification where the threshold for defining substantial habitat modification is set at 50% instead of 30%"}
finalModel_age50 <- readRDS(paste0(modelsDirRobustness,"ModelAge50.rds"))

draws <- as_draws_df(finalModel_age50)[,3:15]

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("2k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxLow1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv501","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

plotCols <- c("#D55E00","#D55E00",
              "#D55E00","#D55E00","#D55E00","#D55E00",
              "#F0E442","#F0E442","#E69F00","#E69F00",
              "#009E73","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Variable,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Variable)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Variable,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Variable),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```

```{r figforestpesthigh, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from the full binomial Bayesian hierarchical model relating bumblebee occurrence to land use, landscape habitat, pesticide toxicity and thermal niche properties**. From the model testing robustness to fitting estimates of duration of substantial habitat modification where the threshold for defining substantial habitat modification is set at 50% instead of 30%"}
finalModel_pesthigh <- readRDS(paste0(modelsDirRobustness,"ModelPestHigh.rds"))

draws <- as_draws_df(finalModel_pesthigh)[,3:15]

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("2k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxHigh1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv1","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

plotCols <- c("#D55E00","#D55E00",
              "#D55E00","#D55E00","#D55E00","#D55E00",
              "#F0E442","#F0E442","#E69F00","#E69F00",
              "#009E73","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Variable,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Variable)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Variable,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Variable),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```