---
title: Supplementary Material for 'Bumble bee probability of occurrence responds to interactions between local and landscape land use, climatic niche properties and climate change'
author: "Tim Newbold^1,\\*^, Jeremy Kerr^2^, Peter Soroye^2,&dagger;^, Jessica J. Williams^1,&Dagger;^"
output:
  bookdown::word_document2:
    toc: false
    fig_caption: TRUE
    number_sections: no
always_allow_html: true
csl: ecology-letters.csl
bibliography: data_references.bib
nocite: '@*'
---

^1^Centre for Biodiversity and Environment Research, Department of Genetics, Evolution and Environment, University College London, London, UK

^2^Department of Biology, University of Ottawa, Ottawa, Ontario, Canada

^&dagger;^Current address: Wildlife Conservation Society Canada, Toronto, Ontario, Canada

^&Dagger;^Current address: Department of Life Sciences, Silwood Park Campus, Imperial College London, London, UK

^\*^Correspondence: [t.newbold@ucl.ac.uk](mailto:t.newbold@ucl.ac.uk)

<!-- SETUP: Global options -->
```{r,echo=FALSE,results=FALSE}
knitr::opts_chunk$set
options(scipen=20)
set.seed(1234)
```

<!-- SETUP: read data and models -->
```{r,echo=FALSE,results=FALSE,warning=FALSE}
# Load required packages
suppressMessages(suppressWarnings(library(terra))) # Version 1.7-71
suppressMessages(suppressWarnings(library(brms))) # Version 2.21.0
suppressMessages(suppressWarnings(library(Hmisc))) # Version 5.1-1
suppressMessages(suppressWarnings(library(ggplot2))) # Version 3.5.1
suppressMessages(suppressWarnings(library(ggstance))) # Version 0.3.7
suppressMessages(suppressWarnings(library(cowplot))) # Version 1.1.3
suppressMessages(suppressWarnings(library(sf))) # Version 1.0-19
suppressMessages(suppressWarnings(library(snow))) # Version 0.4-4
suppressMessages(suppressWarnings(library(parallel))) # Version 4.3.1
suppressMessages(suppressWarnings(library(spdep))) # Version 1.2-8
suppressMessages(suppressWarnings(library(dplyr))) # Version 1.1.3
suppressMessages(suppressWarnings(library(tidyr))) # Version 1.3.1

# Set input directories
dataDir <- "2_PrepareDiversityData/"
modelsDir <- "3_RunModels/"
modelsDirRobustness <- "6_ModelRobustnessChecks/"

# Load the compiled species-level data
diversity <- readRDS(paste0(dataDir,"diversity_data.rds"))

allData <- readRDS(paste0(modelsDir,"ModelData.Rds"))

sites <- readRDS(paste0(modelsDir,"AnalysisSites.rds"))

# Set input directory
inDir <- "3_RunModels/"

# Log-transform elevation, which is very skewed
diversity$LogElevation <- log(diversity$Elevation+2)

# Log-transform pesticide application, which is also very skewed
diversity$LogPestToxLow <- log(diversity$PestToxLow+1)

diversity$UI <- paste(diversity$LandUse,diversity$Use_intensity,sep="_")
diversity$UI[grepl("Natural",diversity$UI)] <- "Natural"
diversity$UI[grepl("NA",diversity$UI)] <- NA
diversity$UI[grepl("Cannot decide",diversity$UI)] <- NA
diversity$UI <- factor(diversity$UI)
diversity$UI <- relevel(diversity$UI,ref="Natural")

# Select relevant columns
modelData <- diversity[,c('occur','LandUse','TEI_BL',
                          'TEI_delta','LogElevation','SS','SSBS',
                          'Taxon_name_entered','Longitude','Latitude',
                          'Best_guess_binomial','Country',
                          'LogPestToxLow',
                          'NaturalHabitat2k','AgeConv')]

# Remove rows with any NA values
modelData <- na.omit(modelData)

# Load the final model
finalModel <- readRDS(paste0(inDir,"FinalModelBRMS.rds"))
```

<!-- Run spatial autocorrelation tests -->
```{r, echo=FALSE,results=FALSE,eval=TRUE,warning=FALSE}

modelData <- droplevels(modelData)

siteAvgResids <- data.frame(resids=tapply(X = residuals(finalModel)[,1],
                                          INDEX = droplevels(modelData)$SSBS,
                                          FUN = mean))
siteAvgResids$SSBS <- row.names(siteAvgResids)
row.names(siteAvgResids) <- NULL
siteAvgResids$Longitude <- modelData$Longitude[
  match(siteAvgResids$SSBS,modelData$SSBS)]
siteAvgResids$Latitude <- modelData$Latitude[
  match(siteAvgResids$SSBS,modelData$SSBS)]
siteAvgResids$SS <- modelData$SS[
  match(siteAvgResids$SSBS,modelData$SSBS)]

moran <- as.data.frame(t(sapply(X = as.character(unique(siteAvgResids$SS)),FUN = function(ss){
  
  data.sub <- droplevels(siteAvgResids[siteAvgResids$SS==ss,])
  
  ds.nb<-try(dnearneigh(cbind(data.sub$Longitude,data.sub$Latitude),
                            d1=0.00000001,d2=10),silent=TRUE)
  ds.listw<-try(nb2listw(ds.nb),silent=TRUE)
  mt<-tryCatch(moran.test(data.sub$resids,ds.listw),silent=TRUE,error=function(e) e,
               warning=function(w) w)
  
  if(class(mt)[1]=="htest"){
        if ((!is.na(mt$statistic))){
          return(data.frame(SS=ss,i=mt$statistic,p=mt$p.value))
        } else {
          return(data.frame(SS=ss,i=NA,p=NA))
        }
    
  } else {
    return(data.frame(SS=ss,i=NA,p=NA))
  }
  
})))

moran$SS <- as.character(moran$SS)
moran$i <- as.numeric(moran$i)
moran$p <- as.numeric(moran$p)

```

\newpage

# Sites in Analysis

<br />

<!-- PLOTTING: Plot map of sites included in analysis -->
```{r figsitesmap, echo=FALSE, warning=FALSE, fig.width=16.5/2.54, fig.height=12/2.54,dpi=1200, out.width="100%", fig.cap="**Location of sites used in the analysis**. Base map shows the UN sub-regions from which bumble bee data were obtained: Northern America, Central America, Northern Europe, Western Europe and Southern Europe. The 13 countries sampled were: Canada, the United States, Belgium, Estonia, France, Germany, the Republic of Ireland, Italy, the Netherlands, Serbia, Sweden, Switzerland and the United Kingdom. Note that Mexico is classed within Central America in the UN sub-region scheme. Map is plotted with geographical coordinates using the WGS 1984 datum"}

dataDir <- "0_data/"

# Create basemap for North American and Western European sub-regions
baseMap <- vect("0_data/UN_subregion.shp")
baseMap <- terra::subset(baseMap,baseMap$SUBREGION %in% c(21,13,154,155,39))
baseMap <- st_as_sf(baseMap)

# Create map of the sites sampled in PREDICTS
sitesMap <- vect(x = sites,geom=c("Longitude","Latitude"),
                 crs = '+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0')

sitesMap <- st_as_sf(sitesMap)

# Plot base map and overlaid with sampled sites
mapPreds <- ggplot() + geom_sf(data = baseMap,fill = "#ffffff") + 
  geom_sf(data=sitesMap,colour = "#cc0000") +
  scale_x_continuous(limits = c(-180,60)) + 
  theme_classic() + 
  theme(axis.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.text = element_text(size=10))

mapPreds
```

\newpage

# Species in Analysis

```{r, echo=FALSE, warning=FALSE,results=TRUE,message=FALSE}

species.table <- tibble::as_tibble(unique(modelData$Best_guess_binomial))
names(species.table) <- "Species"

kableExtra::kable(x = species.table,caption = "List of all species included in the analysis")

```
\newpage

```{r, echo=FALSE,warning=FALSE,results=FALSE,fig.width=16.5/2.54,fig.height=12/2.54,out.width="100%",dpi=1200,fig.cap="**Representation of species with different threat statuses in the model**. For European bumble bee species, for which all species have been assigned a Red List status (https://portals.iucn.org/library/sites/library/files/documents/RL-4-019.pdf), the distribution of sampled and unsampled species among different levels of extinction risk (LC = Least Concern, NT = Near Threatened, VU = Vulnerable, EN = Endangered, CR = Critically Endangered, DD = Data Deficient). Reveals a bias in sampling towards species with lower levels of assessed extinction risk."}

sp.threat.samples <- read.csv(paste0(dataDir,"BumblebeeThreatStatus.csv"))

threat.sample.counts <- suppressMessages(suppressWarnings(
  sp.threat.samples %>% 
  group_by(Status, Data) %>% 
  summarise(n = n()) %>%
  ungroup %>%
  complete(Status, Data, fill = list(n = 0))))

threat.sample.counts$Status <- factor(threat.sample.counts$Status,
                                      levels=c("LC","NT","VU","EN","CR","DD"))

stackbar <- ggplot(data = threat.sample.counts, 
                   mapping = aes(x = Status,y = n,fill = Data)) + 
  geom_bar(stat='identity', position = 'stack') + 
  scale_y_continuous(name = 'Number of species') + 
  scale_fill_manual(values = c("#BF2C23","#2F67B1")) + 
  labs(fill = "Species Sampled") + 
  theme_classic()

stackbar

```

\newpage

# Distributions of Explanatory Variables

<br />

```{r explVarsHists,echo=FALSE,warning=FALSE,results=FALSE,fig.width=16.5/2.54,fig.height=21.5/2.54,out.width="100%",dpi=1200,fig.cap="**Distributions of the records included in the analysis along the gradients of the explanatory variables**. Histograms are plotted separately for natural habitats (blue colour) and human-modified land uses (red colour). Black points indicate the position of records within studies that showed significant residual spatial autocorrelation in the final model"}

# In order to show position along explanatory variable gradients of studies with
# significant spatial autocorrelation, set colour for points so that studies with
# significant spatial autocorrelation plot as black, and all other points as fully
# transparent white (i.e., invisible)
modelData$PointCol <- "#ffffff00"
modelData$PointCol[modelData$SS %in% moran$SS[which(moran$p<0.05)]] <- "#000000"

histNatHab <- ggplot(data = modelData) + 
  geom_histogram(mapping = aes(x = NaturalHabitat2k/10,fill=LandUse),
                 bins=10,position="dodge") + 
  geom_point(mapping = aes(x = NaturalHabitat2k/10,y = 7300),col = modelData$PointCol) + 
  scale_x_continuous(name = "Natural Habitat (%)") + 
  scale_y_continuous(name = "Frequency") + 
  scale_fill_manual(values = c("#2F67B1","#BF2C23")) + 
  theme(axis.title = element_text(size = 12),axis.text = element_text(size = 10)) + 
  theme_classic()

histAge <- ggplot(data = modelData) + 
  geom_histogram(mapping = aes(x = AgeConv,fill=LandUse),
                 bins=10,position="dodge") + 
  geom_point(mapping = aes(x = AgeConv,y = 7000),col = modelData$PointCol) + 
  scale_x_continuous(name = "Duration of landscape conversion (yrs)") + 
  scale_y_continuous(name = "Frequency") + 
  scale_fill_manual(values = c("#2F67B1","#BF2C23")) + 
  theme(axis.title = element_text(size = 12),axis.text = element_text(size = 10)) +
  theme_classic()

histPest <- ggplot(data = modelData) + 
  geom_histogram(mapping = aes(x = exp(LogPestToxLow)-1,fill=LandUse),
                 bins=10,position="dodge") + 
  geom_point(mapping = aes(x = exp(LogPestToxLow)-1,y = 5800),col = modelData$PointCol) + 
  scale_x_continuous(name = "Pesticide toxicity") + 
  scale_y_continuous(name = "Frequency") + 
  scale_fill_manual(values = c("#2F67B1","#BF2C23")) + 
  theme(axis.title = element_text(size = 12),axis.text = element_text(size = 10)) +
  theme_classic()

histTEIBL <- ggplot(data = modelData) + 
  geom_histogram(mapping = aes(x = TEI_BL,fill=LandUse),
                 bins=10,position="dodge") + 
  geom_point(mapping = aes(x = TEI_BL,y = 5800),col = modelData$PointCol) + 
  scale_x_continuous(name = "Baseline thermal position") + 
  scale_y_continuous(name = "Frequency") + 
  scale_fill_manual(values = c("#2F67B1","#BF2C23")) + 
  theme(axis.title = element_text(size = 12),axis.text = element_text(size = 10)) +
  theme_classic()

histTEIDelta <- ggplot(data = modelData) + 
  geom_histogram(mapping = aes(x = TEI_delta,fill=LandUse),
                 bins=10,position="dodge") + 
  geom_point(mapping = aes(x = TEI_delta,y = 4600),col = modelData$PointCol) + 
  scale_x_continuous(name = "\u0394 thermal position") + 
  scale_y_continuous(name = "Frequency") + 
  scale_fill_manual(values = c("#2F67B1","#BF2C23")) + 
  theme(axis.title = element_text(size = 12),axis.text = element_text(size = 10)) +
  theme_classic()

plots <- plot_grid(histNatHab,histAge,histPest,histTEIBL,histTEIDelta,
                   nrow = 3,ncol = 2)

plots
```

\newpage

# Model Diagnostic Checks

<br />    

```{r diagnostic setup,echo=FALSE,results=FALSE,warning=FALSE}

vars <- variables(finalModel)

vars <-  vars[grepl("^b_",vars)]

```

<!-- PLOTTING: model diagnostic checks -->
```{r diagnostic1,echo=FALSE,warning=FALSE, fig.width=16.5/2.54, fig.height=22.5/2.54,dpi=1200,out.width="100%"}

plot(finalModel,variable=vars[1:4])

```

```{r diagnostic2,echo=FALSE,warning=FALSE, fig.width=16.5/2.54, fig.height=22.5/2.54,dpi=1200,out.width="100%"}

plot(finalModel,variable=vars[5:8])

```

```{r diagnostic3,echo=FALSE,warning=FALSE, fig.width=16.5/2.54, fig.height=22.5/2.54,dpi=1200,out.width="100%"}

plot(finalModel,variable=vars[9:12])

```

```{r diagn4,echo=FALSE,results=FALSE,warning=FALSE,fig.width=16.5/2.54,fig.height=16.875/2.54,dpi=1200,out.width="100%",fig.cap="**Test**"}

plot(finalModel,variable=vars[13:15])

```
<br />

```{r ppcheck,echo=FALSE,results=FALSE,warning=FALSE,fig.width=8.25/2.54,fig.height=8/2.54,dpi=1200,out.width="50%",fig.cap="**Density plot comparing observed presences/absences with model-predicted values drawn from the posterior distribution**. Figure is produced directly by the *pp_check* function of the brms R package Version 2.21.0"}
pp_check(finalModel,ndraws=100)
```

<br /> 

```{r spatautocorrel,echo=FALSE,results=FALSE,warning=FALSE, fig.width=8.5, fig.height=7,out.width="50%", fig.cap="**Distribution of P values from a series of Moran's tests for spatial autocorrelation in the residuals associated with each individual study**. For each study, we calculated the average residual for each sampled site, and then ran a Moran's test for spatial autocorrelation in these average residuals. Moran's tests were performed using the spdep R package Version 1.2-8"}

par(las=1)
par(tck=-0.01)
par(mgp=c(1.8,0.2,0))
par(mar=c(2.8,2.8,0.2,0.2))
par(cex=2)
par(cex.axis=1.3)
par(cex.lab=1.3)
par(cex.text=1.3)
hist(x = moran$p,breaks = seq(from=0,to=1,by=0.05),main=NA,xlab="Moran's Test P Value")
box(bty="l")
abline(v=0.05,col="#ff0000",lwd=3)

text(x=0.07,y=8,labels=paste0("Significant \nspatial \nautocorrelation \nin ",round(length(which(moran$p<0.05))/length(which(!is.na(moran$p)))*100,1),"% \nof studies"),
     pos=4)

```

\newpage

# Model Robustness Checks

<br />


```{r figforestnathab1, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from the full binomial Bayesian hierarchical model relating bumble bee occurrence to land use, landscape habitat, pesticide toxicity and thermal niche properties**. From the model testing robustness to fitting estimates of landscape natural habitat at 1-km instead of 2-km spatial grain. Points represent median coefficient estimates, thick horizontal bars the 67% credible intervals, and thin bars the 95% credible intervals. Effects whose 95% credible intervals do not cross zero are interpreted as being 'significant'. Coefficient estimates are coloured by variable grouping: purple - land use; green - landscape natural habitat; light orange - pesticide toxicity; yellow - duration of substantial landscape habitat modification to human uses; dark orange - realized thermal niche position and effect on this of climate change. Text in parentheses refers to whether the relationship is for sites with natural local habitat or with human land use. For the baseline thermal niche position, the Baseline Thermal Position and Baseline Thermal Position^2 coefficients refer to the linear and quadratic components of the polynomial relationship, respectively, combined to describe the curivlinear relationship shown in Figure 2 in the main text."}
finalModel_nathab1 <- readRDS(paste0(modelsDirRobustness,"ModelNatHab1k.rds"))

draws <- as_draws_df(finalModel_nathab1)[,3:15]

draws$`b_LandUseHuman:polyNaturalHabitat1k1` <-
  draws$`b_LandUseHuman:polyNaturalHabitat1k1` + draws$b_polyNaturalHabitat1k1
draws$`b_LandUseHuman:polyLogPestToxLow1` <- 
  draws$`b_LandUseHuman:polyLogPestToxLow1` + draws$b_polyLogPestToxLow1
draws$`b_LandUseHuman:polyAgeConv1` <- 
  draws$`b_LandUseHuman:polyAgeConv1` + draws$b_polyAgeConv1
draws$`b_LandUseHuman:polyTEI_BL21` <- 
  draws$`b_LandUseHuman:polyTEI_BL21` + draws$b_polyTEI_BL21
draws$`b_LandUseHuman:polyTEI_BL22` <- 
  draws$`b_LandUseHuman:polyTEI_BL22` + draws$b_polyTEI_BL22
draws$`b_LandUseHuman:polyTEI_delta1` <- 
  draws$`b_LandUseHuman:polyTEI_delta1` + draws$b_polyTEI_delta1

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("1k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxLow1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv1","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

draw.summary$Label <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Human Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat (Natural)",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat (Human)",
                                    "PesticideToxicity" = "Pesticide Toxicity (Natural)",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity (Human)",
                                    "ConversionDuration" = "Landscape Conversion Duration (Natural)",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration (Human)",
                                    "ThermalPosition_Baseline" = "Baseline Thermal Position (Natural)",
                                    "ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Natural)",
                                    "ThermalPosition_Delta" = "Thermal Position Change (Natural)",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Baseline Thermal Position (Human)",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Human)",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Position Change (Human)")

draw.summary$Group <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat",
                                    "PesticideToxicity" = "Pesticide Toxicity",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity",
                                    "ConversionDuration" = "Landscape Conversion Duration",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration",
                                    "ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "ThermalPosition_Delta" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Niche Properties")

plotCols <- c("#D55E00","#F0E442","#E69F00","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Label,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Group)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Label,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Group),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  guides(colour = guide_legend(nrow = 3,reverse = TRUE,
                               override.aes = aes(size=0.5))) + 
  theme_classic() +
  labs(colour = "Variable Group") + 
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```

```{r figforestnathab5, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from the full binomial Bayesian hierarchical model relating bumble bee occurrence to land use, landscape habitat, pesticide toxicity and thermal niche properties**. From the model testing robustness to fitting estimates of landscape natural habitat at 5-km instead of 2-km spatial grain. Points represent median coefficient estimates, thick horizontal bars the 67% credible intervals, and thin bars the 95% credible intervals. Effects whose 95% credible intervals do not cross zero are interpreted as being 'significant'. Coefficient estimates are coloured by variable grouping: purple - land use; green - landscape natural habitat; light orange - pesticide toxicity; yellow - duration of substantial landscape habitat modification to human uses; dark orange - realized thermal niche position and effect on this of climate change. Text in parentheses refers to whether the relationship is for sites with natural local habitat or with human land use. For the baseline thermal niche position, the Baseline Thermal Position and Baseline Thermal Position^2 coefficients refer to the linear and quadratic components of the polynomial relationship, respectively, combined to describe the curivlinear relationship shown in Figure 2 in the main text."}
finalModel_nathab5 <- readRDS(paste0(modelsDirRobustness,"ModelNatHab5k.rds"))

draws <- as_draws_df(finalModel_nathab5)[,3:15]

draws$`b_LandUseHuman:polyNaturalHabitat5k1` <-
  draws$`b_LandUseHuman:polyNaturalHabitat5k1` + draws$b_polyNaturalHabitat5k1
draws$`b_LandUseHuman:polyLogPestToxLow1` <- 
  draws$`b_LandUseHuman:polyLogPestToxLow1` + draws$b_polyLogPestToxLow1
draws$`b_LandUseHuman:polyAgeConv1` <- 
  draws$`b_LandUseHuman:polyAgeConv1` + draws$b_polyAgeConv1
draws$`b_LandUseHuman:polyTEI_BL21` <- 
  draws$`b_LandUseHuman:polyTEI_BL21` + draws$b_polyTEI_BL21
draws$`b_LandUseHuman:polyTEI_BL22` <- 
  draws$`b_LandUseHuman:polyTEI_BL22` + draws$b_polyTEI_BL22
draws$`b_LandUseHuman:polyTEI_delta1` <- 
  draws$`b_LandUseHuman:polyTEI_delta1` + draws$b_polyTEI_delta1

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("5k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxLow1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv1","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

draw.summary$Label <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Human Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat (Natural)",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat (Human)",
                                    "PesticideToxicity" = "Pesticide Toxicity (Natural)",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity (Human)",
                                    "ConversionDuration" = "Landscape Conversion Duration (Natural)",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration (Human)",
                                    "ThermalPosition_Baseline" = "Baseline Thermal Position (Natural)",
                                    "ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Natural)",
                                    "ThermalPosition_Delta" = "Thermal Position Change (Natural)",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Baseline Thermal Position (Human)",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Human)",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Position Change (Human)")

draw.summary$Group <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat",
                                    "PesticideToxicity" = "Pesticide Toxicity",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity",
                                    "ConversionDuration" = "Landscape Conversion Duration",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration",
                                    "ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "ThermalPosition_Delta" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Niche Properties")

plotCols <- c("#D55E00","#F0E442","#E69F00","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Label,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Group)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Label,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Group),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  guides(colour = guide_legend(nrow = 3,reverse = TRUE,
                               override.aes = aes(size=0.5))) + 
  theme_classic() +
  labs(colour = "Variable Group") + 
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```

```{r figforestage10, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from the full binomial Bayesian hierarchical model relating bumble bee occurrence to land use, landscape habitat, pesticide toxicity and thermal niche properties**. From the model testing robustness to fitting estimates of duration of substantial habitat modification where the threshold for defining substantial habitat modification is set at 10% instead of 30%. Points represent median coefficient estimates, thick horizontal bars the 67% credible intervals, and thin bars the 95% credible intervals. Effects whose 95% credible intervals do not cross zero are interpreted as being 'significant'. Coefficient estimates are coloured by variable grouping: purple - land use; green - landscape natural habitat; light orange - pesticide toxicity; yellow - duration of substantial landscape habitat modification to human uses; dark orange - realized thermal niche position and effect on this of climate change. Text in parentheses refers to whether the relationship is for sites with natural local habitat or with human land use. For the baseline thermal niche position, the Baseline Thermal Position and Baseline Thermal Position^2 coefficients refer to the linear and quadratic components of the polynomial relationship, respectively, combined to describe the curivlinear relationship shown in Figure 2 in the main text."}
finalModel_age10 <- readRDS(paste0(modelsDirRobustness,"ModelAge10.rds"))

draws <- as_draws_df(finalModel_age10)[,3:15]

draws$`b_LandUseHuman:polyNaturalHabitat2k1` <-
  draws$`b_LandUseHuman:polyNaturalHabitat2k1` + draws$b_polyNaturalHabitat2k1
draws$`b_LandUseHuman:polyLogPestToxLow1` <- 
  draws$`b_LandUseHuman:polyLogPestToxLow1` + draws$b_polyLogPestToxLow1
draws$`b_LandUseHuman:polyAgeConv101` <- 
  draws$`b_LandUseHuman:polyAgeConv101` + draws$b_polyAgeConv101
draws$`b_LandUseHuman:polyTEI_BL21` <- 
  draws$`b_LandUseHuman:polyTEI_BL21` + draws$b_polyTEI_BL21
draws$`b_LandUseHuman:polyTEI_BL22` <- 
  draws$`b_LandUseHuman:polyTEI_BL22` + draws$b_polyTEI_BL22
draws$`b_LandUseHuman:polyTEI_delta1` <- 
  draws$`b_LandUseHuman:polyTEI_delta1` + draws$b_polyTEI_delta1

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("2k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxLow1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv101","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

draw.summary$Label <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Human Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat (Natural)",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat (Human)",
                                    "PesticideToxicity" = "Pesticide Toxicity (Natural)",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity (Human)",
                                    "ConversionDuration" = "Landscape Conversion Duration (Natural)",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration (Human)",
                                    "ThermalPosition_Baseline" = "Baseline Thermal Position (Natural)",
                                    "ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Natural)",
                                    "ThermalPosition_Delta" = "Thermal Position Change (Natural)",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Baseline Thermal Position (Human)",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Human)",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Position Change (Human)")

draw.summary$Group <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat",
                                    "PesticideToxicity" = "Pesticide Toxicity",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity",
                                    "ConversionDuration" = "Landscape Conversion Duration",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration",
                                    "ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "ThermalPosition_Delta" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Niche Properties")

plotCols <- c("#D55E00","#F0E442","#E69F00","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Label,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Group)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Label,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Group),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  guides(colour = guide_legend(nrow = 3,reverse = TRUE,
                               override.aes = aes(size=0.5))) + 
  theme_classic() +
  labs(colour = "Variable Group") + 
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```

```{r figforestage50, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from the full binomial Bayesian hierarchical model relating bumble bee occurrence to land use, landscape habitat, pesticide toxicity and thermal niche properties**. From the model testing robustness to fitting estimates of duration of substantial habitat modification where the threshold for defining substantial habitat modification is set at 50% instead of 30%. Points represent median coefficient estimates, thick horizontal bars the 67% credible intervals, and thin bars the 95% credible intervals. Effects whose 95% credible intervals do not cross zero are interpreted as being 'significant'. Coefficient estimates are coloured by variable grouping: purple - land use; green - landscape natural habitat; light orange - pesticide toxicity; yellow - duration of substantial landscape habitat modification to human uses; dark orange - realized thermal niche position and effect on this of climate change. Text in parentheses refers to whether the relationship is for sites with natural local habitat or with human land use. For the baseline thermal niche position, the Baseline Thermal Position and Baseline Thermal Position^2 coefficients refer to the linear and quadratic components of the polynomial relationship, respectively, combined to describe the curivlinear relationship shown in Figure 2 in the main text."}
finalModel_age50 <- readRDS(paste0(modelsDirRobustness,"ModelAge50.rds"))

draws <- as_draws_df(finalModel_age50)[,3:15]

draws$`b_LandUseHuman:polyNaturalHabitat2k1` <-
  draws$`b_LandUseHuman:polyNaturalHabitat2k1` + draws$b_polyNaturalHabitat2k1
draws$`b_LandUseHuman:polyLogPestToxLow1` <- 
  draws$`b_LandUseHuman:polyLogPestToxLow1` + draws$b_polyLogPestToxLow1
draws$`b_LandUseHuman:polyAgeConv501` <- 
  draws$`b_LandUseHuman:polyAgeConv501` + draws$b_polyAgeConv501
draws$`b_LandUseHuman:polyTEI_BL21` <- 
  draws$`b_LandUseHuman:polyTEI_BL21` + draws$b_polyTEI_BL21
draws$`b_LandUseHuman:polyTEI_BL22` <- 
  draws$`b_LandUseHuman:polyTEI_BL22` + draws$b_polyTEI_BL22
draws$`b_LandUseHuman:polyTEI_delta1` <- 
  draws$`b_LandUseHuman:polyTEI_delta1` + draws$b_polyTEI_delta1

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("2k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxLow1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv501","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

draw.summary$Label <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Human Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat (Natural)",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat (Human)",
                                    "PesticideToxicity" = "Pesticide Toxicity (Natural)",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity (Human)",
                                    "ConversionDuration" = "Landscape Conversion Duration (Natural)",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration (Human)",
                                    "ThermalPosition_Baseline" = "Baseline Thermal Position (Natural)",
                                    "ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Natural)",
                                    "ThermalPosition_Delta" = "Thermal Position Change (Natural)",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Baseline Thermal Position (Human)",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Human)",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Position Change (Human)")

draw.summary$Group <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat",
                                    "PesticideToxicity" = "Pesticide Toxicity",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity",
                                    "ConversionDuration" = "Landscape Conversion Duration",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration",
                                    "ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "ThermalPosition_Delta" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Niche Properties")

plotCols <- c("#D55E00","#F0E442","#E69F00","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Label,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Group)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Label,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Group),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  guides(colour = guide_legend(nrow = 3,reverse = TRUE,
                               override.aes = aes(size=0.5))) + 
  theme_classic() +
  labs(colour = "Variable Group") + 
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```


```{r figforestpesthigh, echo=FALSE, warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from the full binomial Bayesian hierarchical model relating bumble bee occurrence to land use, landscape habitat, pesticide toxicity and thermal niche properties**. From the model testing robustness to using high instead of low estimates of pesticide application density. Points represent median coefficient estimates, thick horizontal bars the 67% credible intervals, and thin bars the 95% credible intervals. Effects whose 95% credible intervals do not cross zero are interpreted as being 'significant'. Coefficient estimates are coloured by variable grouping: purple - land use; green - landscape natural habitat; light orange - pesticide toxicity; yellow - duration of substantial landscape habitat modification to human uses; dark orange - realized thermal niche position and effect on this of climate change. Text in parentheses refers to whether the relationship is for sites with natural local habitat or with human land use. For the baseline thermal niche position, the Baseline Thermal Position and Baseline Thermal Position^2 coefficients refer to the linear and quadratic components of the polynomial relationship, respectively, combined to describe the curivlinear relationship shown in Figure 2 in the main text."}
finalModel_pesthigh <- readRDS(paste0(modelsDirRobustness,"ModelPestHigh.rds"))

draws <- as_draws_df(finalModel_pesthigh)[,3:15]

draws$`b_LandUseHuman:polyNaturalHabitat2k1` <-
  draws$`b_LandUseHuman:polyNaturalHabitat2k1` + draws$b_polyNaturalHabitat2k1
draws$`b_LandUseHuman:polyLogPestToxHigh1` <- 
  draws$`b_LandUseHuman:polyLogPestToxHigh1` + draws$b_polyLogPestToxHigh1
draws$`b_LandUseHuman:polyAgeConv1` <- 
  draws$`b_LandUseHuman:polyAgeConv1` + draws$b_polyAgeConv1
draws$`b_LandUseHuman:polyTEI_BL21` <- 
  draws$`b_LandUseHuman:polyTEI_BL21` + draws$b_polyTEI_BL21
draws$`b_LandUseHuman:polyTEI_BL22` <- 
  draws$`b_LandUseHuman:polyTEI_BL22` + draws$b_polyTEI_BL22
draws$`b_LandUseHuman:polyTEI_delta1` <- 
  draws$`b_LandUseHuman:polyTEI_delta1` + draws$b_polyTEI_delta1

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("2k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxHigh1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv1","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

draw.summary$Label <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Human Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat (Natural)",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat (Human)",
                                    "PesticideToxicity" = "Pesticide Toxicity (Natural)",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity (Human)",
                                    "ConversionDuration" = "Landscape Conversion Duration (Natural)",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration (Human)",
                                    "ThermalPosition_Baseline" = "Baseline Thermal Position (Natural)",
                                    "ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Natural)",
                                    "ThermalPosition_Delta" = "Thermal Position Change (Natural)",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Baseline Thermal Position (Human)",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Human)",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Position Change (Human)")

draw.summary$Group <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat",
                                    "PesticideToxicity" = "Pesticide Toxicity",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity",
                                    "ConversionDuration" = "Landscape Conversion Duration",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration",
                                    "ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "ThermalPosition_Delta" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Niche Properties")

plotCols <- c("#D55E00","#F0E442","#E69F00","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Label,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Group)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Label,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Group),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  guides(colour = guide_legend(nrow = 3,reverse = TRUE,
                               override.aes = aes(size=0.5))) + 
  theme_classic() +
  labs(colour = "Variable Group") + 
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```


```{r, figrelabundmodel, echo=FALSE,warning=FALSE, fig.width=17.5/2.54, fig.height=12/2.54,out.width="100%", dpi=1200, fig.cap="**Forest plot of coefficient estimates from a zero-inflated negative binomial model relating bumble bee relative abundance to land use, landscape habitat, pesticide toxicity and thermal niche properties**. Points represent median coefficient estimates, thick horizontal bars the 67% credible intervals, and thin bars the 95% credible intervals. Effects whose 95% credible intervals do not cross zero are interpreted as being 'significant'. Coefficient estimates are coloured by variable grouping: purple - land use; green - landscape natural habitat; light orange - pesticide toxicity; yellow - duration of substantial landscape habitat modification to human uses; dark orange - realized thermal niche position and effect on this of climate change. Text in parentheses refers to whether the relationship is for sites with natural local habitat or with human land use. For the baseline thermal niche position, the Baseline Thermal Position and Baseline Thermal Position^2 coefficients refer to the linear and quadratic components of the polynomial relationship, respectively, combined to describe the curivlinear relationship shown in Figure 2 in the main text."}

finalModel_relabund <- readRDS(paste0(modelsDirRobustness,"ModelZINBAbund.rds"))

draws <- as_draws_df(finalModel_relabund)[,3:15]

draws$`b_LandUseHuman:polyNaturalHabitat2k1` <-
  draws$`b_LandUseHuman:polyNaturalHabitat2k1` + draws$b_polyNaturalHabitat2k1
draws$`b_LandUseHuman:polyLogPestToxLow1` <- 
  draws$`b_LandUseHuman:polyLogPestToxLow1` + draws$b_polyLogPestToxLow1
draws$`b_LandUseHuman:polyAgeConv1` <- 
  draws$`b_LandUseHuman:polyAgeConv1` + draws$b_polyAgeConv1
draws$`b_LandUseHuman:polyTEI_BL21` <- 
  draws$`b_LandUseHuman:polyTEI_BL21` + draws$b_polyTEI_BL21
draws$`b_LandUseHuman:polyTEI_BL22` <- 
  draws$`b_LandUseHuman:polyTEI_BL22` + draws$b_polyTEI_BL22
draws$`b_LandUseHuman:polyTEI_delta1` <- 
  draws$`b_LandUseHuman:polyTEI_delta1` + draws$b_polyTEI_delta1

draw.means <- apply(X = draws,MARGIN = 2,FUN = mean)
draw.lower.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.025)
draw.upper.95 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.975)
draw.lower.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.16667)
draw.upper.67 <- apply(X = draws,MARGIN = 2,FUN = quantile,probs = 0.83333)

draw.summary <- data.frame(Variable = names(draw.means),
                           Mean = draw.means,
                           Lower95 = draw.lower.95,
                           Upper95 = draw.upper.95,
                           Lower67 = draw.lower.67,
                           Upper67 = draw.upper.67)

draw.summary$Variable <- gsub("b_","",draw.summary$Variable)
draw.summary$Variable <- gsub("poly","",draw.summary$Variable)
draw.summary$Variable <- gsub("2k1","",draw.summary$Variable)
draw.summary$Variable <- gsub("LogPestToxLow1","PesticideToxicity",draw.summary$Variable)
draw.summary$Variable <- gsub("AgeConv1","ConversionDuration",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL21","ThermalPosition_Baseline",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_BL22","ThermalPosition_Baseline2",draw.summary$Variable)
draw.summary$Variable <- gsub("TEI_delta1","ThermalPosition_Delta",draw.summary$Variable)
# draw.summary$Variable <- gsub("LogElevation1","Elevation",draw.summary$Variable)

draw.summary$Variable <- factor(draw.summary$Variable,
                                levels = c("LandUseHuman:ThermalPosition_Delta",
                                           "ThermalPosition_Delta",
                                           "LandUseHuman:ThermalPosition_Baseline2",
                                           "LandUseHuman:ThermalPosition_Baseline",
                                           "ThermalPosition_Baseline2",
                                           "ThermalPosition_Baseline",
                                           "LandUseHuman:ConversionDuration",
                                           "ConversionDuration",
                                           "LandUseHuman:PesticideToxicity",
                                           "PesticideToxicity",
                                           "LandUseHuman:NaturalHabitat",
                                           "NaturalHabitat",
                                           "LandUseHuman"))

draw.summary$Label <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Human Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat (Natural)",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat (Human)",
                                    "PesticideToxicity" = "Pesticide Toxicity (Natural)",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity (Human)",
                                    "ConversionDuration" = "Landscape Conversion Duration (Natural)",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration (Human)",
                                    "ThermalPosition_Baseline" = "Baseline Thermal Position (Natural)",
                                    "ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Natural)",
                                    "ThermalPosition_Delta" = "Thermal Position Change (Natural)",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Baseline Thermal Position (Human)",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Baseline Thermal Position^2 (Human)",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Position Change (Human)")

draw.summary$Group <- dplyr::recode(draw.summary$Variable,
                                    "LandUseHuman" = "Land Use",
                                    "NaturalHabitat" = "Landscape Natural Habitat",
                                    "LandUseHuman:NaturalHabitat" = "Landscape Natural Habitat",
                                    "PesticideToxicity" = "Pesticide Toxicity",
                                    "LandUseHuman:PesticideToxicity" = "Pesticide Toxicity",
                                    "ConversionDuration" = "Landscape Conversion Duration",
                                    "LandUseHuman:ConversionDuration" = "Landscape Conversion Duration",
                                    "ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "ThermalPosition_Delta" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Baseline2" = "Thermal Niche Properties",
                                    "LandUseHuman:ThermalPosition_Delta" = "Thermal Niche Properties")

plotCols <- c("#D55E00","#F0E442","#E69F00","#009E73","#CC79A7")

forestPlot <- ggplot() + geom_linerangeh(mapping = aes(y=draw.summary$Label,
                                        xmin = draw.summary$Lower95, 
                                        xmax = draw.summary$Upper95,
                                        colour = draw.summary$Group)) + 
  geom_pointrangeh(mapping = aes(y = draw.summary$Label,
                                 x = draw.summary$Mean,
                                 xmin = draw.summary$Lower67,
                                 xmax = draw.summary$Upper67,
                                 colour = draw.summary$Group),
                   fatten=1.8,size=1) +
  scale_colour_manual(values = plotCols) +
  geom_vline(mapping = aes(xintercept = 0)) +
  labs(x = "Effect size",y = element_blank()) +
  guides(colour = guide_legend(nrow = 3,reverse = TRUE,
                               override.aes = aes(size=0.5))) + 
  theme_classic() +
  labs(colour = "Variable Group") + 
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
  
forestPlot
```

\newpage

# References to Original Data Papers in PREDICTS Database
